---
title: "Data preparation"
---

```{r setup}
source(here::here("R", "00_packages.R"))
source(here::here("R", "01_paths.R"))
source(here::here("R", "02_read_clean_data.R"))
source(here::here("R", "03_filters.R"))
```

## Read and inspect the raw data

```{r read-raw}
dat_raw <- read_proceed()

cat("Rows:", nrow(dat_raw), "\n")
cat("Columns:", ncol(dat_raw), "\n")
```

### Column names

```{r column-names}
names(dat_raw)
```

### Quick look at key variables

```{r glimpse}
dplyr::glimpse(dat_raw)
```

## Initial cleaning

Convert grouping and categorical variables to factors.

```{r clean}
dat_clean <- clean_proceed(dat_raw)
```

## Inclusion criteria

### Require lnM input variables

Each contrast must have valid means, standard deviations, and sample sizes in both groups.

```{r require-lnm}
n_before <- nrow(dat_clean)
dat_filtered <- require_lnm_vars(dat_clean)
n_after  <- nrow(dat_filtered)

cat("Excluded (missing lnM inputs):", n_before - n_after, "\n")
cat("Retained:", n_after, "\n")
```

### Generation filter

The primary analysis applies a **generations ≤ 300** filter where generation-time information is available. Contrasts with missing generation time are retained.

```{r gen-filter}
n_before <- nrow(dat_filtered)
dat_filtered <- filter_generations(dat_filtered, max_gen = 300)
n_after  <- nrow(dat_filtered)

cat("Excluded (generations > 300):", n_before - n_after, "\n")
cat("Retained:", n_after, "\n")
```

### Design coverage

The analysis includes both allochronic (before–after) and synchronic (disturbed–reference) contrasts unless a specific chapter restricts to one design. The `design` variable is retained as a moderator.

```{r design-counts}
dat_filtered |>
  dplyr::count(design, sort = TRUE)
```

## Derived variables

```{r derived}
dat_lnm_input <- add_derived_vars(dat_filtered)

dplyr::glimpse(
  dplyr::select(dat_lnm_input, n_total, log10_years, log10_generations)
)
```

## Quick dataset summary

```{r summary}
dat_lnm_input |>
  dplyr::summarise(
    k              = dplyr::n(),
    n_studies      = dplyr::n_distinct(ref_id),
    n_systems      = dplyr::n_distinct(sys_id),
    n_species      = dplyr::n_distinct(sp_ncbi),
    n_allochronic  = sum(design == "Allochronic",  na.rm = TRUE),
    n_synchronic   = sum(design == "Synchronic",   na.rm = TRUE),
    pct_has_gen    = round(100 * mean(!is.na(generations)), 1),
    pct_has_years  = round(100 * mean(!is.na(years) & years > 0), 1)
  )
```

## Save cleaned dataset

```{r save}
saveRDS(
  dat_lnm_input,
  here::here("outputs", "data_clean", "proceed_clean_filtered.rds")
)
cat("Saved: outputs/data_clean/proceed_clean_filtered.rds\n")
```
