---
title: "lnM effect sizes (SAFE)"
---

```{r setup}
source(here::here("R", "00_packages.R"))
source(here::here("R", "01_paths.R"))
source(here::here("R", "04_lnm_safe.R"))
source(here::here("R", "15_render_guard.R"))   # read-only: never recompute

# Set to TRUE to recompute (slow; ~hours for large datasets)
recompute_effect_sizes <- FALSE
```

## Load cleaned data

```{r load-clean}
dat_lnm_input <- readRDS(
  here::here("outputs", "data_clean", "proceed_clean_filtered.rds")
)

cat("Contrasts available for lnM:", nrow(dat_lnm_input), "\n")
```

## SAFE lnM estimator

The SAFE estimator (`orchaRd:::.safe_lnM_indep()`) uses Monte Carlo sampling to compute a bias-corrected estimate of $\ln M$ and its sampling variance for each independent-groups contrast. For each contrast, a large number of draws are taken from the posterior predictive distribution of the two group means, standard deviations, and sample sizes; the estimator retains only draws that satisfy validity constraints (e.g., positive standard deviations) and computes $\ln M$ for each.

This is computationally intensive but avoids the negative-bias that affects naive $\ln M$ estimates from small samples.

## Compute or read SAFE lnM estimates

```{r compute-lnm}
dat_es <- compute_or_read_lnm_safe(
  dat_lnm_input,
  out_path   = here::here("outputs", "effect_sizes", "proceed_lnm_safe.rds"),
  diag_path  = here::here("outputs", "tables", "safe_lnm_diagnostics.csv"),
  recompute  = recompute_effect_sizes
)

cat("Effect sizes retained:", nrow(dat_es), "\n")
```

## Effect-size summary

```{r es-summary}
dat_es |>
  dplyr::summarise(
    k           = dplyr::n(),
    min_yi      = round(min(yi_lnM_safe), 3),
    median_yi   = round(median(yi_lnM_safe), 3),
    max_yi      = round(max(yi_lnM_safe), 3),
    min_vi      = round(min(vi_lnM_safe), 6),
    median_vi   = round(median(vi_lnM_safe), 6),
    max_vi      = round(max(vi_lnM_safe), 6),
    n_status_ok = sum(status == "ok", na.rm = TRUE),
    n_failed    = sum(status != "ok", na.rm = TRUE)
  )
```

## SAFE diagnostics

Check the Monte Carlo convergence (draws kept vs. total).

```{r safe-diag}
safe_diag <- readr::read_csv(
  here::here("outputs", "tables", "safe_lnm_diagnostics.csv"),
  show_col_types = FALSE
)

safe_diag |>
  dplyr::summarise(
    n_ok           = sum(status == "ok"),
    n_not_ok       = sum(status != "ok"),
    min_pct_kept   = round(min(draws_kept / draws_total * 100), 1),
    median_pct_kept = round(median(draws_kept / draws_total * 100), 1),
    max_attempts   = max(attempts)
  )
```

## Identifier management

The dataset preserves two IDs:

- **`es_id_db`**: the original PROCEED effect-size identifier (for tracing back to the database).
- **`es_id_model`**: a sequential integer factor created *after* filtering to finite, positive $v_i$, used as the row/column index of the sampling-variance matrix $V$.

```{r id-check}
dat_es |>
  dplyr::select(es_id_db, es_id_model) |>
  head(6)
```

## Sampling-variance matrix

The diagonal covariance matrix $V$ is used in the brms model via `(1 | gr(es_id_model, cov = V))`. The corresponding standard deviation prior is set to `constant(1)`.

```{r vcv-dims}
cat("Effect-size rows (= dim of V):", nrow(dat_es), "\n")
```
