---
title: "Dependency structure"
---

```{r setup}
source(here::here("R", "00_packages.R"))
source(here::here("R", "01_paths.R"))
```

## Load data

```{r load}
dat_es <- readRDS(here::here("outputs", "effect_sizes", "proceed_lnm_safe.rds"))
```

## Why dependency matters

Multiple contrasts in a meta-analysis often share a reference (same paper), a study system (same population or location), or a species. Ignoring this non-independence inflates effective sample size and narrows credible intervals inappropriately. Random effects in the location–scale model account for three sources of dependency:

1. **Reference (`ref_id`)** — contrasts from the same paper share design choices, sampling effort, and measurement protocol.
2. **Phylogenetic relatedness (`sp_ncbi`, via A)** — species that share evolutionary history have correlated trait values.
3. **Sampling variance (`es_id_model`, via V)** — each contrast has a known estimation error (from SAFE) that is treated as a fixed component rather than estimated.

`sys_id` (study system) is correlated with both `ref_id` and `sp_ncbi` in these data, which makes it hard to identify a separate system-level variance component. It is therefore reserved for sensitivity analyses.

## Nesting diagnostic helper

```{r nesting-fun}
check_nesting <- function(data, a, b) {
  tmp <- data |>
    dplyr::count(.data[[a]], .data[[b]], name = "n") |>
    dplyr::count(.data[[a]], name = "n_unique_b")

  tibble::tibble(
    factor_a              = a,
    factor_b              = b,
    n_levels_a            = dplyr::n_distinct(data[[a]]),
    n_levels_b            = dplyr::n_distinct(data[[b]]),
    min_unique_b_per_a    = min(tmp$n_unique_b),
    median_unique_b_per_a = stats::median(tmp$n_unique_b),
    max_unique_b_per_a    = max(tmp$n_unique_b),
    prop_a_single_b       = mean(tmp$n_unique_b == 1),
    a_nested_in_b         = all(tmp$n_unique_b == 1)
  )
}
```

## Build diagnostic dataset

```{r diag-data}
dat_re_diag <- dat_es |>
  dplyr::transmute(
    es_id_db = factor(es_id_db),
    ref_id   = factor(ref_id),
    sys_id   = factor(sys_id),
    sp_ncbi  = factor(sp_ncbi)
  ) |>
  tidyr::drop_na() |>
  droplevels()

cat("Rows in diagnostic dataset:", nrow(dat_re_diag), "\n")
```

## Nesting table

If `a_nested_in_b == TRUE`, every level of A belongs to exactly one level of B — A is fully nested within B. If `prop_a_single_b` is high, most levels of A belong to a single level of B.

```{r nesting-table}
group_vars <- c("es_id_db", "ref_id", "sys_id", "sp_ncbi")

nesting_tbl <- purrr::map_dfr(group_vars, function(a) {
  purrr::map_dfr(setdiff(group_vars, a), function(b) {
    check_nesting(dat_re_diag, a, b)
  })
}) |>
  dplyr::arrange(dplyr::desc(a_nested_in_b), factor_a, factor_b)

nesting_tbl
```

## Nesting matrix (max unique B per A level)

A value of 1 means A is fully nested in B.

```{r nesting-matrix}
nesting_tbl |>
  dplyr::select(factor_a, factor_b, max_unique_b_per_a) |>
  tidyr::pivot_wider(names_from = factor_b, values_from = max_unique_b_per_a)
```

## Support summaries

### Systems per species

```{r sp-sys}
species_support <- dat_re_diag |>
  dplyr::count(sp_ncbi, sys_id) |>
  dplyr::count(sp_ncbi, name = "n_sys_per_species") |>
  dplyr::summarise(
    n_species                  = dplyr::n(),
    prop_species_with_1_system = mean(n_sys_per_species == 1),
    median_systems_per_species = stats::median(n_sys_per_species),
    max_systems_per_species    = max(n_sys_per_species)
  )

species_support
```

### Systems per reference

```{r ref-sys}
ref_support <- dat_re_diag |>
  dplyr::count(ref_id, sys_id) |>
  dplyr::count(ref_id, name = "n_sys_per_ref") |>
  dplyr::summarise(
    n_refs                  = dplyr::n(),
    prop_refs_with_1_system = mean(n_sys_per_ref == 1),
    median_systems_per_ref  = stats::median(n_sys_per_ref),
    max_systems_per_ref     = max(n_sys_per_ref)
  )

ref_support
```

### Systems with multiple species

```{r sys-multi-sp}
dat_re_diag |>
  dplyr::count(sys_id, sp_ncbi) |>
  dplyr::count(sys_id, name = "n_species_in_sys") |>
  dplyr::filter(n_species_in_sys > 1) |>
  dplyr::arrange(dplyr::desc(n_species_in_sys))
```

### Systems with multiple references

```{r sys-multi-ref}
dat_re_diag |>
  dplyr::count(sys_id, ref_id) |>
  dplyr::count(sys_id, name = "n_refs_in_sys") |>
  dplyr::filter(n_refs_in_sys > 1) |>
  dplyr::arrange(dplyr::desc(n_refs_in_sys))
```

## Support summary table

```{r support-table}
make_link_summary <- function(data, grouping_factor, linked_factor, relationship_label) {
  data |>
    dplyr::distinct(.data[[grouping_factor]], .data[[linked_factor]]) |>
    dplyr::count(.data[[grouping_factor]], name = "n_matches") |>
    dplyr::summarise(
      Group                      = grouping_factor,
      Relationship               = relationship_label,
      `Groups with only 1 match` = sum(n_matches == 1),
      `Total groups`             = dplyr::n(),
      `Only 1 match (%)`         = round(100 * mean(n_matches == 1), 1)
    )
}

support_table <- dplyr::bind_rows(
  make_link_summary(dat_re_diag, "sp_ncbi", "sys_id",  "systems per species"),
  make_link_summary(dat_re_diag, "sys_id",  "sp_ncbi", "species per system"),
  make_link_summary(dat_re_diag, "ref_id",  "sys_id",  "systems per reference"),
  make_link_summary(dat_re_diag, "sys_id",  "ref_id",  "references per system")
)

support_table
```

## Species coverage for phylogenetic random effect

Before fitting any model it is useful to know how many rows will be affected by the phylogenetic alignment step. Two sources of row loss are distinguished:

1. **`sp_ncbi` is `NA` or blank** — the contrast has no species information at all. These rows are always excluded from any species-level random effect.
2. **Valid `sp_ncbi` not matched in the Open Tree of Life** — the species name exists in the data but `rotl` could not retrieve it during tree construction. Under the default `fill_unmatched = TRUE` behaviour these species are retained in the model and assigned zero phylogenetic correlation with all other species (a "star" contribution). They are **not** silently dropped.

```{r phylo-coverage}
source(here::here("R", "05_phylogeny.R"))

A_full   <- tryCatch(readRDS(here::here("outputs", "phylogeny", "proceed_A_matrix.rds")),
                     error = function(e) NULL)
name_map <- tryCatch(readRDS(here::here("outputs", "phylogeny", "proceed_name_map.rds")),
                     error = function(e) NULL)

if (!is.null(A_full) && !is.null(name_map)) {
  dat_coverage <- apply_phylo_name_map(dat_es, name_map)

  sp_raw       <- as.character(dat_es$sp_ncbi)
  sp_canonical <- as.character(dat_coverage$sp_ncbi_canonical)

  n_na_raw       <- sum(is.na(sp_raw) | sp_raw == "")
  n_named        <- sum(!is.na(sp_raw) & nzchar(sp_raw))
  n_resolved     <- sum(!is.na(sp_canonical) & nzchar(sp_canonical))
  n_unresolved   <- n_named - n_resolved

  sp_uniq_can    <- unique(sp_canonical[!is.na(sp_canonical) & nzchar(sp_canonical)])
  sp_in_tree     <- sp_uniq_can[sp_uniq_can %in% rownames(A_full)]
  sp_star        <- setdiff(sp_uniq_can, rownames(A_full))

  tibble::tibble(
    Category = c(
      "Total contrasts",
      "sp_ncbi is NA / blank (no species info)",
      "Named species — resolved to canonical name",
      "Named species — unresolved after 4-stage cascade",
      "Unique canonical species in phylogenetic tree",
      "Unique canonical species assigned star (zero) correlation"
    ),
    N = c(
      nrow(dat_es),
      n_na_raw,
      n_resolved,
      n_unresolved,
      length(sp_in_tree),
      length(sp_star)
    )
  )
} else {
  cat("Phylogeny / name map not yet built. Run chapter 05 first.\n")
}
```

## Random-effects rationale

The nesting diagnostics above reveal:

- `es_id_db` is the finest-grained identifier (one row per contrast) and is nested within all higher-level groupings.
- Most studies (`ref_id`) contribute data from one system, but a meaningful fraction contribute from multiple systems.
- Most species (`sp_ncbi`) appear in one system, but some span multiple systems and references.
- Systems can span multiple references and species, creating a crossed (not purely hierarchical) structure.

This crossed structure means that `ref_id` and `sp_ncbi` (via phylogeny) can both be estimated in the same model. Adding `sys_id` as a *third* group-level predictor in this crossed design risks near-collinearity with `ref_id` and `sp_ncbi`, making variance components weakly identified. For this reason, `sys_id` is included in sensitivity analyses only.

## Save dependency tables

```{r save}
readr::write_csv(
  nesting_tbl,
  here::here("outputs", "tables", "dependency_nesting_table.csv")
)

readr::write_csv(
  support_table,
  here::here("outputs", "tables", "dependency_support_summary.csv")
)

cat("Dependency tables saved.\n")
```
