---
title: "Location–scale model grid"
---

```{r setup}
source(here::here("R", "00_packages.R"))
source(here::here("R", "01_paths.R"))
source(here::here("R", "05_phylogeny.R"))
source(here::here("R", "06_model_registry.R"))
source(here::here("R", "07_model_formulas.R"))
source(here::here("R", "08_fit_or_read_model.R"))
source(here::here("R", "09_model_summaries.R"))
source(here::here("R", "10_model_diagnostics.R"))
source(here::here("R", "15_render_guard.R"))   # read-only: never refit / rebuild phylogeny

# Change to TRUE only when you want to refit all models from scratch
refit_models <- FALSE

# The grid status + combined effect tables are pre-computed (precompute_grid_summaries.R)
# into outputs/tables/grid_summary_cache.rds so the book never loads a 400 MB model.
# Set TRUE only to regenerate them by loading every model (slow).
rebuild_grid <- FALSE
```

## Overview

This chapter fits (or reads) one independent location–scale meta-regression for each moderator in the registry. No multi-moderator model is fitted. The registry is defined in `R/06_model_registry.R`.

```{r show-registry}
moderator_grid
```

## Load data and phylogeny

```{r load-data}
dat_es <- readRDS(here::here("outputs", "effect_sizes", "proceed_lnm_safe.rds"))
cat("Total contrasts:", nrow(dat_es), "\n")
```

```{r load-phylogeny}
# build_or_read_phylogeny() returns list(A = ..., name_map = ...)
# It uses prepR4pcm for name normalisation + 4-stage reconciliation before rotl.
phylo_out <- tryCatch({
  A_path       <- here::here("outputs", "phylogeny", "proceed_A_matrix.rds")
  name_map_path <- here::here("outputs", "phylogeny", "proceed_name_map.rds")

  if (file.exists(A_path) && file.exists(name_map_path)) {
    list(A        = readRDS(A_path),
         name_map = readRDS(name_map_path))
  } else {
    message("No cached phylogeny — building via prepR4pcm + rotl ...")
    build_or_read_phylogeny(
      species_vec = unique(as.character(dat_es$sp_ncbi)),
      recompute   = FALSE
    )
  }
}, error = function(e) {
  message("Phylogeny load/build failed: ", conditionMessage(e))
  NULL
})

A_full   <- phylo_out$A
name_map <- phylo_out$name_map

if (!is.null(A_full)) {
  cat("Phylogenetic matrix:", nrow(A_full), "×", ncol(A_full), "\n")
} else {
  message("Phylogenetic matrix unavailable — models will omit the phylogenetic term.")
}
```

```{r apply-name-map}
# Apply the prepR4pcm name map to dat_es ONCE here, adding sp_ncbi_canonical.
# prepare_phylo_and_data() will use this column to match rows to A.
# Rows whose sp_ncbi had no canonical match get NA in sp_ncbi_canonical and
# are excluded from the phylogenetic term (but not from other random effects).
if (!is.null(name_map)) {
  dat_es <- apply_phylo_name_map(dat_es, name_map)
  n_resolved <- sum(!is.na(dat_es$sp_ncbi_canonical))
  n_na_orig  <- sum(is.na(dat_es$sp_ncbi) | dat_es$sp_ncbi == "")
  n_unresolved <- sum(
    !is.na(dat_es$sp_ncbi) & dat_es$sp_ncbi != "" &
    is.na(dat_es$sp_ncbi_canonical)
  )
  cat("sp_ncbi_canonical coverage:\n")
  cat("  Resolved to canonical name :", n_resolved, "\n")
  cat("  Original sp_ncbi was NA/blank:", n_na_orig, "\n")
  cat("  Named but unresolved        :", n_unresolved, "\n")
} else {
  # No name map — fall back to using sp_ncbi directly
  dat_es$sp_ncbi_canonical <- dat_es$sp_ncbi
  message("No name map available; using sp_ncbi as-is for phylogeny matching.")
}
```

## Verify moderators exist in the data

```{r check-moderators}
grid_status <- moderator_grid |>
  dplyr::mutate(
    in_data = moderator %in% names(dat_es),
    n_nonmissing = purrr::map_int(moderator, function(m) {
      if (m %in% names(dat_es)) sum(!is.na(dat_es[[m]])) else 0L
    })
  )

grid_status
```

## Fit or read models

The loop below iterates over every row of the registry. For each moderator it:

1. Checks the moderator exists and has enough data.
2. Drops rows missing that moderator.
3. Rebuilds the sequential `es_id_model` and sampling-variance matrix `V`.
4. Subsets the phylogenetic matrix `A` to the species present.
5. Builds the formula and priors.
6. Calls `fit_or_read_model()`, which reads the cached `.rds` if it exists.
7. Extracts summaries and diagnostics.

```{r fit-grid, eval=rebuild_grid, results='hide'}
# Heavy path (rebuild_grid = TRUE only): load every model and extract summaries.
# Normal renders skip this and read the precomputed cache below.
all_fe    <- list()
all_diag  <- list()
status_rows <- list()

for (i in seq_len(nrow(moderator_grid))) {
  row        <- moderator_grid[i, ]
  mod_id     <- row$model_id
  moderator  <- row$moderator
  mod_label  <- row$label
  mod_type   <- row$type
  model_file <- paste0(mod_id, "_ls_", moderator)

  message("\n--- ", mod_id, ": ", moderator, " ---")

  # ── 1. Check moderator exists ──────────────────────────────────────────────
  if (!moderator %in% names(dat_es)) {
    status_rows[[i]] <- tibble::tibble(
      model_id = mod_id, moderator = moderator, label = mod_label,
      type = mod_type, n_rows = 0L, n_levels = NA_integer_,
      n_missing_excluded = NA_integer_,
      model_path = NA_character_, status = "skipped: variable absent",
      notes = "Moderator not found in dataset.",
      max_rhat = NA_real_, min_bulk_ess = NA_real_, min_tail_ess = NA_real_,
      n_divergent = NA_integer_, max_treedepth_hits = NA_integer_
    )
    next
  }

  # ── 2. Subset data for this moderator ─────────────────────────────────────
  n_before     <- nrow(dat_es)
  dat_model    <- dat_es[!is.na(dat_es[[moderator]]), ]
  n_missing_ex <- n_before - nrow(dat_model)

  if (mod_type == "categorical") {
    dat_model[[moderator]] <- droplevels(factor(dat_model[[moderator]]))
  }

  # Check sample size per level for categorical moderators
  n_levels <- NA_integer_
  level_counts <- NULL
  if (mod_type == "categorical") {
    level_counts <- table(dat_model[[moderator]])
    n_levels     <- length(level_counts)
    sparse       <- level_counts[level_counts < 5]
    if (length(sparse) > 0) {
      message("Sparse levels (<5 obs) in ", moderator, ": ",
              paste(names(sparse), collapse = ", "))
    }
    if (all(level_counts < 3)) {
      status_rows[[i]] <- tibble::tibble(
        model_id = mod_id, moderator = moderator, label = mod_label,
        type = mod_type, n_rows = nrow(dat_model), n_levels = n_levels,
        n_missing_excluded = n_missing_ex,
        model_path = NA_character_, status = "skipped: too sparse",
        notes = "All levels have fewer than 3 observations.",
        max_rhat = NA_real_, min_bulk_ess = NA_real_, min_tail_ess = NA_real_,
        n_divergent = NA_integer_, max_treedepth_hits = NA_integer_
      )
      next
    }
  }

  if (nrow(dat_model) < 10) {
    status_rows[[i]] <- tibble::tibble(
      model_id = mod_id, moderator = moderator, label = mod_label,
      type = mod_type, n_rows = nrow(dat_model), n_levels = n_levels,
      n_missing_excluded = n_missing_ex,
      model_path = NA_character_, status = "skipped: insufficient data",
      notes = paste("Only", nrow(dat_model), "rows after missingness exclusion."),
      max_rhat = NA_real_, min_bulk_ess = NA_real_, min_tail_ess = NA_real_,
      n_divergent = NA_integer_, max_treedepth_hits = NA_integer_
    )
    next
  }

  # ── 3 & 4. Align data to phylogeny, build V ─────────────────────────────────
  # prepare_phylo_and_data() subsets A to species present in this model's data,
  # filters dat_model to the intersection (so every sp_ncbi level is in A),
  # rebuilds es_id_model sequentially, and constructs the diagonal V matrix.
  dat_model <- dat_model |>
    dplyr::mutate(es_id_model = factor(seq_len(dplyr::n())))

  phylo     <- prepare_phylo_and_data(dat_model, A_full, label = mod_id)
  dat_model <- phylo$dat_model
  A_mod     <- phylo$A_mod
  V         <- phylo$V
  has_phylo <- phylo$has_phylo

  # ── 5. Build formula and priors ───────────────────────────────────────────
  formula <- build_ls_formula(moderator, has_phylogeny = has_phylo)
  priors  <- build_ls_priors(formula, dat_model, V, A = A_mod)

  # ── 5b. Verify constant(1) is correctly set before any fitting ────────────
  prior_ok <- verify_esid_prior(priors, model_name = mod_id)
  if (!prior_ok) {
    status_rows[[i]] <- tibble::tibble(
      model_id = mod_id, moderator = moderator, label = mod_label,
      type = mod_type, n_rows = nrow(dat_model), n_levels = n_levels,
      n_missing_excluded = n_missing_ex,
      model_path = NA_character_, status = "skipped: prior check failed",
      notes = "constant(1) was not confirmed at coef=Intercept level for es_id_model.",
      max_rhat = NA_real_, min_bulk_ess = NA_real_, min_tail_ess = NA_real_,
      n_divergent = NA_integer_, max_treedepth_hits = NA_integer_
    )
    next
  }

  # ── 6. Fit or read ─────────────────────────────────────────────────────────
  model_path_full <- file.path(
    here::here("outputs", "models"),
    paste0(model_file, ".rds")
  )

  fit <- tryCatch({
    fit_or_read_model(
      model_name = model_file,
      fit_fun    = function() {
        fit_ls_model(dat_model, formula, priors,
                     V = V, A = A_mod,
                     mcmc_args = default_mcmc_args)
      },
      model_dir  = here::here("outputs", "models"),
      refit      = refit_models
    )
  }, error = function(e) {
    message("Model fitting failed for ", moderator, ": ", conditionMessage(e))
    NULL
  })

  # ── 7. Extract summaries and diagnostics ──────────────────────────────────
  if (is.null(fit)) {
    status_rows[[i]] <- tibble::tibble(
      model_id = mod_id, moderator = moderator, label = mod_label,
      type = mod_type, n_rows = nrow(dat_model), n_levels = n_levels,
      n_missing_excluded = n_missing_ex,
      model_path = model_path_full, status = "failed",
      notes = "Model returned NULL (fitting error).",
      max_rhat = NA_real_, min_bulk_ess = NA_real_, min_tail_ess = NA_real_,
      n_divergent = NA_integer_, max_treedepth_hits = NA_integer_
    )
    next
  }

  fe   <- extract_fixed_effects(fit, mod_id, moderator)
  diag <- tryCatch(
    extract_diagnostics(fit, mod_id, moderator),
    error = function(e) {
      message("Diagnostics failed for ", moderator, ": ", conditionMessage(e))
      tibble::tibble(
        model_id = mod_id, moderator = moderator,
        max_rhat = NA_real_, min_bulk_ess = NA_real_,
        min_tail_ess = NA_real_, n_divergent = NA_integer_,
        max_treedepth_hits = NA_integer_
      )
    }
  )

  all_fe[[mod_id]]   <- split_location_scale(fe)
  all_diag[[mod_id]] <- diag

  status_rows[[i]] <- tibble::tibble(
    model_id           = mod_id,
    moderator          = moderator,
    label              = mod_label,
    type               = mod_type,
    n_rows             = nrow(dat_model),
    n_levels           = n_levels,
    n_missing_excluded = n_missing_ex,
    model_path         = model_path_full,
    status             = "fitted",
    notes              = "",
    max_rhat           = diag$max_rhat,
    min_bulk_ess       = diag$min_bulk_ess,
    min_tail_ess       = diag$min_tail_ess,
    n_divergent        = diag$n_divergent,
    max_treedepth_hits = diag$max_treedepth_hits
  )
}
```

## Assemble grid summaries

```{r grid-assemble}
grid_cache <- here::here("outputs", "tables", "grid_summary_cache.rds")

if (rebuild_grid) {
  # Persist what the loop above produced (also refreshes the CSVs + cache).
  model_status <- dplyr::bind_rows(status_rows)
  readr::write_csv(model_status, here::here("outputs", "tables", "model_grid_status.csv"))
  summ     <- save_combined_summaries(all_fe,  tables_dir = here::here("outputs", "tables"))
  diag_tbl <- save_combined_diagnostics(all_diag, tables_dir = here::here("outputs", "tables"))
  loc_effects <- summ$location; scl_effects <- summ$scale
  saveRDS(list(status = model_status, location = loc_effects,
               scale = scl_effects, diagnostics = diag_tbl), grid_cache)
} else {
  # Normal render: read the small precomputed cache — no 400 MB model is loaded.
  pace_need(grid_cache, "model grid summary cache")
  g <- readRDS(grid_cache)
  model_status <- g$status; loc_effects <- g$location
  scl_effects  <- g$scale;  diag_tbl    <- g$diagnostics
}

model_status <- model_status |>
  dplyr::mutate(
    status = dplyr::case_when(
      model_id == "m06" ~ "withheld: replacement model pending",
      model_id == "m09" ~ "withheld: replacement model pending",
      TRUE ~ status
    ),
    notes = dplyr::case_when(
      model_id == "m06" ~ "Original taxonomic model had poor convergence and sparse-level geometry; do not interpret this row as a current main result until the replacement model is finalized.",
      model_id == "m09" ~ "Original data-type model had poor scale-parameter diagnostics and sparse levels; do not interpret this row as a current main result until the replacement model is finalized.",
      TRUE ~ notes
    )
  )
```

## Model status table

The taxonomic-group (`m06`) and data-type (`m09`) result pages are temporarily withheld from the rendered book navigation. The original fits had known convergence or sparse-level problems, and the replacement models are being finalized before these moderators are presented as main results.

```{r status-table}
model_status
```

## Convergence overview

```{r convergence}
model_status |>
  dplyr::filter(status == "fitted") |>
  dplyr::select(model_id, moderator, max_rhat, min_bulk_ess, n_divergent) |>
  dplyr::mutate(
    rhat_ok = dplyr::if_else(max_rhat <= 1.01, "ok", "warn"),
    ess_ok  = dplyr::if_else(min_bulk_ess >= 400, "ok", "warn"),
    div_ok  = dplyr::if_else(n_divergent < 10, "ok", "warn")
  )
```
