---
title: "Results by moderator"
---

```{r setup}
source(here::here("R", "00_packages.R"))
source(here::here("R", "01_paths.R"))
source(here::here("R", "06_model_registry.R"))
```

This chapter presents the results of each location–scale model. No models are fitted here. All tables are read from `outputs/tables/`.

## Interpretation guide

**Location effects** are estimated on the $\ln M$ scale. A positive estimate means the moderator level is associated with a *larger* average magnitude of phenotypic divergence. A negative estimate means smaller average divergence. The reference (baseline) level is shown in the model intercept.

**Scale effects** are estimated on the $\log(\sigma)$ scale. A positive estimate means the moderator level is associated with *greater residual heterogeneity* in lnM — divergence is more unpredictable or variable across contrasts of that type. A negative estimate means *lower residual heterogeneity* — divergence is more consistent and predictable.

**Avoid binary significance language.** The credible intervals quantify uncertainty; the biological question is whether the estimate and its interval are practically meaningful.

## Load result tables

```{r load-tables}
status_file <- here::here("outputs", "tables", "model_grid_status.csv")
loc_file    <- here::here("outputs", "tables", "location_effects_all_moderators.csv")
scl_file    <- here::here("outputs", "tables", "scale_effects_all_moderators.csv")
diag_file   <- here::here("outputs", "tables", "model_diagnostics_all_moderators.csv")

if (file.exists(status_file)) {
  model_status <- readr::read_csv(status_file, show_col_types = FALSE)
} else {
  stop("model_grid_status.csv not found. Run chapter 05 first.")
}

loc_effects  <- if (file.exists(loc_file))  readr::read_csv(loc_file,  show_col_types = FALSE) else NULL
scl_effects  <- if (file.exists(scl_file))  readr::read_csv(scl_file,  show_col_types = FALSE) else NULL
diag_all     <- if (file.exists(diag_file)) readr::read_csv(diag_file, show_col_types = FALSE) else NULL
```

## Model grid status

```{r status}
model_status |>
  dplyr::select(model_id, moderator, label, type, n_rows, status, max_rhat, n_divergent)
```

## Results loop

```{r results-loop, results='asis'}
for (i in seq_len(nrow(moderator_grid))) {
  row       <- moderator_grid[i, ]
  mod_id    <- row$model_id
  moderator <- row$moderator
  label     <- row$label

  cat("\n\n### ", mod_id, " — ", label, " {.unnumbered}\n\n", sep = "")

  # Status
  st <- model_status[model_status$model_id == mod_id, ]
  if (nrow(st) == 0) {
    cat("*Status: no information available.*\n\n")
    next
  }

  cat("**Status:** ", st$status, "\n\n")
  cat("**N contrasts:** ", st$n_rows, "  |  ",
      "**N excluded (missing):** ", st$n_missing_excluded, "\n\n")

  if (st$status != "fitted") {
    cat("*Model was not fitted. Reason: ", st$notes, "*\n\n")
    next
  }

  # Diagnostics
  if (!is.null(diag_all)) {
    d <- diag_all[diag_all$model_id == mod_id, ]
    if (nrow(d) > 0) {
      cat("**Diagnostics:** max Rhat =", round(d$max_rhat, 4),
          "| min bulk ESS =", round(d$min_bulk_ess),
          "| divergent transitions =", d$n_divergent, "\n\n")
    }
  }

  # Location effects
  if (!is.null(loc_effects)) {
    lfe <- loc_effects[loc_effects$model_id == mod_id, ]
    if (nrow(lfe) > 0) {
      cat("**Location effects (mean lnM):**\n\n")
      print(knitr::kable(
        dplyr::select(lfe, term, estimate, est_error, q2_5, q97_5) |>
          dplyr::mutate(dplyr::across(where(is.numeric), ~ round(.x, 3))),
        format = "markdown"
      ))
      cat("\n\n")
    }
  }

  # Scale effects
  if (!is.null(scl_effects)) {
    sfe <- scl_effects[scl_effects$model_id == mod_id, ]
    if (nrow(sfe) > 0) {
      cat("**Scale effects (residual heterogeneity, log sigma):**\n\n")
      print(knitr::kable(
        dplyr::select(sfe, term, estimate, est_error, q2_5, q97_5) |>
          dplyr::mutate(dplyr::across(where(is.numeric), ~ round(.x, 3))),
        format = "markdown"
      ))
      cat("\n\n")
    }
  }
}
```
