---
title: "Reproducibility"
---

## Computational environment {#sec-reproducibility}

```{r setup}
source(here::here("R", "00_packages.R"))
```

### R and package versions

```{r session-info}
sessioninfo::session_info()
```

### CmdStan version

```{r cmdstan-version}
tryCatch(
  cat("CmdStan version:", cmdstanr::cmdstan_version(), "\n"),
  error = function(e) cat("CmdStan not available:", conditionMessage(e), "\n")
)
```

### prepR4pcm version

```{r prepR4pcm-version}
tryCatch(
  cat("prepR4pcm version:",
      as.character(utils::packageVersion("prepR4pcm")), "\n"),
  error = function(e) cat("prepR4pcm not installed:", conditionMessage(e), "\n")
)
```

### Phylogeny reconciliation summary

```{r phylo-recon-summary}
name_map_path <- here::here("outputs", "phylogeny", "proceed_name_map.rds")
if (file.exists(name_map_path)) {
  nm <- readRDS(name_map_path)
  cat("Name-map entries:", nrow(nm), "\n")
  cat("Resolved        :", sum(!is.na(nm$sp_ncbi_canonical)), "\n")
  cat("Unresolved      :", sum( is.na(nm$sp_ncbi_canonical)), "\n")
} else {
  cat("Name map not found — phylogeny not yet built.\n")
}
```

### Git information

```{r git-info}
git_hash <- tryCatch(
  system("git rev-parse HEAD", intern = TRUE),
  error   = function(e) "git not available"
)

git_status <- tryCatch(
  system("git status --short", intern = TRUE),
  error   = function(e) character(0)
)

cat("Commit hash:", git_hash, "\n\n")
if (length(git_status) > 0) {
  cat("Uncommitted changes:\n")
  cat(paste(git_status, collapse = "\n"), "\n")
} else {
  cat("Working tree clean.\n")
}
```

### Random seeds

All SAFE lnM computations use `seed = 123`. All brms model fits use `seed = 123`. These seeds are set in `R/04_lnm_safe.R` and `R/08_fit_or_read_model.R` respectively.

### Path conventions

All paths are constructed with `here::here()` relative to the project root. The project root is determined by the presence of `_quarto.yml` and `Magnitude_pace_modern_life.Rproj`.

```{r here-root}
cat("Project root:", here::here(), "\n")
```

### Model cache inventory

```{r model-cache}
model_files <- list.files(
  here::here("outputs", "models"),
  pattern = "\\.rds$",
  recursive = TRUE,
  full.names = TRUE
)

if (length(model_files) > 0) {
  tibble::tibble(
    path     = model_files,
    filename = basename(model_files),
    size_mb  = round(file.size(model_files) / 1024^2, 2),
    modified = file.mtime(model_files)
  ) |>
    dplyr::arrange(modified)
} else {
  cat("No cached model files found in outputs/models/.\n")
}
```

### Effect-size cache

```{r es-cache}
es_path <- here::here("outputs", "effect_sizes", "proceed_lnm_safe.rds")
if (file.exists(es_path)) {
  cat("Effect-size cache:", es_path, "\n")
  cat("Size:", round(file.size(es_path) / 1024^2, 2), "MB\n")
  cat("Modified:", format(file.mtime(es_path)), "\n")
} else {
  cat("Effect-size cache not found. Run chapter 02.\n")
}
```

### How to render

```r
# Default: reads all caches
quarto::quarto_render()

# Recompute effect sizes (set in 02_lnm_effect_sizes.qmd)
# recompute_effect_sizes <- TRUE

# Refit all models (set in 05_location_scale_model_grid.qmd)
# refit_models <- TRUE
```
