---
title: "Data preparation"
---

```{r setup}
source(here::here("chapters", "Rscripts", "00_packages.R"))
source(here::here("chapters", "Rscripts", "01_paths.R"))
source(here::here("chapters", "Rscripts", "02_read_clean_data.R"))
source(here::here("chapters", "Rscripts", "03_filters.R"))
source(here::here("chapters", "Rscripts", "15_render_guard.R"))   # read-only: never rebuild the database

# The cleaned/filtered contrast dataset is cached as an .rds. The book READS it
# by default. Set to TRUE only to rebuild the database from the raw PROCEED CSV.
rebuild_clean_data <- FALSE

clean_path <- here::here("Rdata", "data_clean", "proceed_clean_filtered.rds")
```

This chapter produces the cleaned, filtered contrast dataset (`proceed_clean_filtered.rds`) used throughout the book. By default it **reads the cached file**; the cleaning pipeline below is documentation and only executes when `rebuild_clean_data <- TRUE`.

## Cleaning & filtering pipeline

The steps applied when (re)building the dataset:

1. `read_proceed()` - read the raw PROCEED v6.2 rates database.
2. `clean_proceed()` - coerce grouping/categorical variables to factors.
3. `require_lnm_vars()` - keep contrasts with the two-group summary statistics required for $\ln M$: means, standard deviations, and sample sizes for both groups.
4. `filter_generations(max_gen = 300)` - retain contrasts with a known, positive elapsed-generation value no greater than 300; missing generation information is excluded because eligibility cannot be established.
5. `add_derived_vars()` - add `n_total`, `log10_years`, `log10_generations`.

```{r rebuild, eval=rebuild_clean_data}
dat_raw       <- read_proceed()
dat_clean     <- clean_proceed(dat_raw)
dat_filtered  <- require_lnm_vars(dat_clean)
dat_filtered  <- filter_generations(dat_filtered, max_gen = 300)
dat_lnm_input <- add_derived_vars(dat_filtered)

saveRDS(dat_lnm_input, clean_path)
cat("Rebuilt and saved cleaned dataset:", nrow(dat_lnm_input), "contrasts\n")
```

## Load the cleaned dataset

```{r load}
if (!rebuild_clean_data) {
  pace_need(clean_path, "cleaned dataset")
  dat_lnm_input <- readRDS(clean_path)
  cat("Read cached cleaned dataset:", nrow(dat_lnm_input), "contrasts\n")
}
```

## Design coverage

```{r design-counts}
dat_lnm_input |>
  dplyr::count(design, sort = TRUE)
```

## Derived variables

```{r derived}
dplyr::glimpse(
  dplyr::select(dat_lnm_input, n_total, log10_years, log10_generations)
)
```

## Dataset summary

```{r summary}
dat_lnm_input |>
  dplyr::summarise(
    k              = dplyr::n(),
    n_studies      = dplyr::n_distinct(ref_id),
    n_systems      = dplyr::n_distinct(sys_id),
    n_species      = dplyr::n_distinct(sp_ncbi),
    n_allochronic  = sum(design == "Allochronic",  na.rm = TRUE),
    n_synchronic   = sum(design == "Synchronic",   na.rm = TRUE),
    pct_has_gen    = round(100 * mean(!is.na(generations)), 1),
    pct_has_years  = round(100 * mean(!is.na(years) & years > 0), 1)
  )
```
