---
title: "lnM effect sizes (SAFE)"
---

```{r setup}
source(here::here("chapters", "Rscripts", "00_packages.R"))
source(here::here("chapters", "Rscripts", "01_paths.R"))
source(here::here("chapters", "Rscripts", "04_lnm_safe.R"))
source(here::here("chapters", "Rscripts", "15_render_guard.R"))   # read-only: never recompute

# Set to TRUE to recompute (slow; ~hours for large datasets)
recompute_effect_sizes <- FALSE
```

## What lnM measures

This chapter computes the effect size used throughout the book: $\ln M$, the log magnitude of separation between two groups. The key idea is to measure how far apart two group means are without retaining the sign of the difference. This is useful here because human disturbance can push traits in different directions across systems, while the biological question is often about the magnitude of phenotypic divergence.

For a two-group comparison with observed summaries $(\bar{X}_1, s_1, n_1)$ and $(\bar{X}_2, s_2, n_2)$, the analytic definition begins with the one-way ANOVA mean squares:

$$
MS_B =
\frac{n_1 n_2}{n_1 + n_2}
(\bar{X}_1 - \bar{X}_2)^2,
$$

$$
MS_W =
\frac{(n_1 - 1)s_1^2 + (n_2 - 1)s_2^2}
{n_1 + n_2 - 2}.
$$

The within-group variance is $s_W^2 = MS_W$. The between-component variance is:

$$
s_B^2 =
\frac{MS_B - MS_W}{n_0},
\qquad
n_0 = \frac{2n_1n_2}{n_1+n_2},
$$

where $n_0$ is the harmonic-mean sample size. The effect size is then:

$$
\ln M = \ln\left(\frac{s_B}{s_W}\right),
$$

where $s_B$ is the square root of the between-component variance and $s_W$ is the pooled within-group standard deviation. The sign of $\ln M$ is therefore not the direction of trait change. Instead:

- $\ln M > 0$ means the estimated between-component standard deviation is larger than the within-group standard deviation.
- $\ln M = 0$ means those two standard deviations are equal.
- $\ln M < 0$ means the estimated between-component standard deviation is smaller than the within-group standard deviation.

Because $s_B^2$ subtracts $MS_W$ from $MS_B$, $\ln M = 0$ is not equivalent to "small" separation. For large, balanced independent groups it corresponds approximately to $d = \sqrt{2}$, or about $1.4$, on the standardized mean-difference scale.

This differs from taking an absolute value of a signed effect size such as $|d|$ or $|\ln RR|$. Absolute-value transformations fold sampling error around zero, which can inflate small and noisy effects. The lnM paper develops $\ln M$ so that magnitude can be analysed directly with standard meta-analytic and meta-regression tools.

## Why SAFE is used here

Analytic $\ln M$ is built from ANOVA mean-square components. That construction is intuitive, but it has two practical problems for large ecological databases:

- Sampling variance can become very large when $MS_B$ is close to $MS_W$.
- The plug-in estimate is undefined when $MS_B \le MS_W$, because $s_B^2$ is then zero or negative.

The SAFE estimator used below follows the paper's simulation-based approach. For each contrast, it repeatedly draws plausible values of the group means and standard deviations from their finite-sample distributions, transforms each valid draw with $MS_B^\ast > MS_W^\ast$ to $\ln M^\ast$, and then summarizes the resulting bootstrap distribution. This gives a finite bias-corrected estimate and sampling variance for contrasts that would otherwise be unstable or undefined.

The book uses the cached SAFE results by default. Recomputing all contrasts is intentionally opt-in because the Monte Carlo step is slow.

## Load cleaned data

```{r load-clean}
dat_lnm_input <- readRDS(
  here::here("Rdata", "data_clean", "proceed_clean_filtered.rds")
)

cat("Contrasts available for lnM:", nrow(dat_lnm_input), "\n")
```

## Compute or read SAFE lnM estimates

```{r compute-lnm}
dat_es <- compute_or_read_lnm_safe(
  dat_lnm_input,
  out_path   = here::here("Rdata", "effect_sizes", "proceed_lnm_safe.rds"),
  diag_path  = here::here("Rdata", "tables", "safe_lnm_diagnostics.csv"),
  recompute  = recompute_effect_sizes
)

cat("Effect sizes retained:", nrow(dat_es), "\n")
```

The resulting dataset carries the SAFE point estimate (`yi_lnM_safe`) and its sampling variance (`vi_lnM_safe`). These two columns become the response and known sampling-error input for the location-scale meta-regression chapters.

## Effect-size summary

```{r es-summary}
dat_es |>
  dplyr::summarise(
    k           = dplyr::n(),
    min_yi      = round(min(yi_lnM_safe), 3),
    median_yi   = round(median(yi_lnM_safe), 3),
    max_yi      = round(max(yi_lnM_safe), 3),
    min_vi      = round(min(vi_lnM_safe), 6),
    median_vi   = round(median(vi_lnM_safe), 6),
    max_vi      = round(max(vi_lnM_safe), 6),
    n_status_ok = sum(status == "ok", na.rm = TRUE),
    n_failed    = sum(status != "ok", na.rm = TRUE)
  )
```

This table is a quick range check, not an inferential result. The median and range of `yi_lnM_safe` describe observed magnitudes on the log scale; the variance columns describe how precisely each contrast estimates its own magnitude.

## SAFE diagnostics

SAFE only summarizes valid simulation draws, so the retained-draw diagnostics are important. A low retained percentage means many simulated draws landed in the region where the between-group component was non-positive; those contrasts are still usable, but their uncertainty should be interpreted with care.

```{r safe-diag}
safe_diag <- readr::read_csv(
  here::here("Rdata", "tables", "safe_lnm_diagnostics.csv"),
  show_col_types = FALSE
)

safe_diag |>
  dplyr::summarise(
    n_ok           = sum(status == "ok"),
    n_not_ok       = sum(status != "ok"),
    min_pct_kept   = round(min(draws_kept / draws_total * 100), 1),
    median_pct_kept = round(median(draws_kept / draws_total * 100), 1),
    max_attempts   = max(attempts)
  )
```

## Identifier management

The dataset preserves two IDs:

- `es_id_db`: the original PROCEED effect-size identifier, kept for tracing estimates back to the database.
- `es_id_model`: a sequential factor created after filtering to finite, positive $v_i$, used as the row/column index of the sampling-variance matrix $V$.

```{r id-check}
dat_es |>
  dplyr::select(es_id_db, es_id_model) |>
  head(6)
```

## Sampling-variance matrix

The diagonal covariance matrix $V$ carries the known SAFE sampling variances into `brms` via `(1 | gr(es_id_model, cov = V))`. The corresponding standard deviation prior is fixed with `constant(1)`, so this term contributes known measurement error rather than estimating a new variance component.

```{r vcv-dims}
cat("Effect-size rows (= dim of V):", nrow(dat_es), "\n")
```
