---
title: "lnM effect sizes (SAFE)"
---

::: {.workflow}
::: {}
**Input**
Clean contrast data from Step 1.
:::
::: {}
**Action**
Calculate SAFE lnM and its sampling variance.
:::
::: {}
**Output**
`Rdata/effect_sizes/proceed_lnm_safe.rds`
:::
::: {}
**Next**
Describe the effect-size dataset.
:::
:::

```{r setup}
source(here::here("Scripts", "00_packages.R"))
source(here::here("Scripts", "01_paths.R"))
source(here::here("Scripts", "04_lnm_safe.R"))
source(here::here("Scripts", "15_render_guard.R"))   # read-only: never recompute

# Set to TRUE to recompute (slow; ~hours for large datasets)
recompute_effect_sizes <- FALSE
```

## Definition of lnM {#sec-lnm-measures}

This chapter computes $\ln M$, the log ratio of the estimated between-group
standard deviation to the pooled within-group standard deviation. The formula
does not retain the direction of the difference between group means.

For a two-group comparison with observed summaries $(\bar{X}_1, s_1, n_1)$ and $(\bar{X}_2, s_2, n_2)$, the analytic definition begins with the one-way ANOVA mean squares:

$$
MS_B =
\frac{n_1 n_2}{n_1 + n_2}
(\bar{X}_1 - \bar{X}_2)^2,
$$

$$
MS_W =
\frac{(n_1 - 1)s_1^2 + (n_2 - 1)s_2^2}
{n_1 + n_2 - 2}.
$$

The within-group variance is $s_W^2 = MS_W$. The between-component variance is:

$$
s_B^2 =
\frac{MS_B - MS_W}{n_0},
\qquad
n_0 = \frac{2n_1n_2}{n_1+n_2},
$$

where $n_0$ is the harmonic-mean sample size. The effect size is then:

$$
\ln M = \ln\left(\frac{s_B}{s_W}\right),
$$

where $s_B$ is the square root of the between-component variance and $s_W$ is the pooled within-group standard deviation. The sign of $\ln M$ is therefore not the direction of trait change. Instead:

- $\ln M > 0$ means the estimated between-component standard deviation is larger than the within-group standard deviation.
- $\ln M = 0$ means the two estimated components have a ratio of one.
- $\ln M < 0$ means the estimated between-component standard deviation is smaller than the within-group standard deviation.

For large, balanced independent groups, $\ln M = 0$ corresponds approximately
to $d = \sqrt{2}$, or about 1.4, on the standardized mean-difference scale.

This calculation differs from taking an absolute value of a signed effect size
such as $|d|$ or $|\ln RR|$. The SAFE code below calculates $\ln M$ directly.

## SAFE calculation

The analytic calculation has two numerical limits:

- Sampling variance can become very large when $MS_B$ is close to $MS_W$.
- The plug-in estimate is undefined when $MS_B \le MS_W$, because $s_B^2$ is then zero or negative.

The SAFE estimator repeatedly draws values of the group means and standard
deviations from their finite-sample distributions. It transforms each draw with
$MS_B^\ast > MS_W^\ast$ to $\ln M^\ast$, then saves the bias-corrected estimate
and sampling variance.

The book uses the cached SAFE results by default. Recomputing all contrasts is opt-in. The Monte Carlo step is slow.

## Load cleaned data

```{r load-clean}
dat_lnm_input <- readRDS(
  here::here("Rdata", "data_clean", "proceed_clean_filtered.rds")
)

cat("Contrasts available for lnM:", nrow(dat_lnm_input), "\n")
```

## Compute or read SAFE lnM estimates

```{r compute-lnm}
dat_es <- compute_or_read_lnm_safe(
  dat_lnm_input,
  out_path   = here::here("Rdata", "effect_sizes", "proceed_lnm_safe.rds"),
  diag_path  = here::here("Rdata", "tables", "safe_lnm_diagnostics.csv"),
  recompute  = recompute_effect_sizes
)

cat("Effect sizes retained:", nrow(dat_es), "\n")
```

The resulting dataset carries the SAFE point estimate (`yi_lnM_safe`) and its sampling variance (`vi_lnM_safe`). These two columns become the response and known sampling-error input for the location-scale meta-regression chapters.

## Effect-size summary

```{r es-summary}
dat_es |>
  dplyr::summarise(
    k           = dplyr::n(),
    min_yi      = round(min(yi_lnM_safe), 3),
    median_yi   = round(median(yi_lnM_safe), 3),
    max_yi      = round(max(yi_lnM_safe), 3),
    min_vi      = round(min(vi_lnM_safe), 6),
    median_vi   = round(median(vi_lnM_safe), 6),
    max_vi      = round(max(vi_lnM_safe), 6),
    n_status_ok = sum(status == "ok", na.rm = TRUE),
    n_failed    = sum(status != "ok", na.rm = TRUE)
  )
```

This table records the median and range of `yi_lnM_safe` and
`vi_lnM_safe`.

## SAFE diagnostics

SAFE summarizes valid simulation draws. The retained percentage records the
share of draws for which the between-group component was positive.

```{r safe-diag}
safe_diag <- readr::read_csv(
  here::here("Rdata", "tables", "safe_lnm_diagnostics.csv"),
  show_col_types = FALSE
)

safe_diag |>
  dplyr::summarise(
    n_ok           = sum(status == "ok"),
    n_not_ok       = sum(status != "ok"),
    min_pct_kept   = round(min(draws_kept / draws_total * 100), 1),
    median_pct_kept = round(median(draws_kept / draws_total * 100), 1),
    max_attempts   = max(attempts)
  )
```

## Identifier management

The dataset preserves two IDs:

- `es_id_db`: the original PROCEED effect-size identifier, kept for tracing estimates back to the database.
- `es_id_model`: a sequential factor created after filtering to finite, positive $v_i$, used as the row/column index of the sampling-variance matrix $V$.

```{r id-check}
dat_es |>
  dplyr::select(es_id_db, es_id_model) |>
  head(6)
```

## Sampling-variance matrix

The diagonal covariance matrix $V$ carries the known SAFE sampling variances into `brms` via `(1 | gr(es_id_model, cov = V))`. The corresponding standard deviation prior is fixed with `constant(1)`, so this term contributes known measurement error rather than estimating a new variance component.

```{r vcv-dims}
cat("Effect-size rows (= dim of V):", nrow(dat_es), "\n")
```
