---
title: "Associations among moderators"
---

```{r setup}
source(here::here("chapters", "Rscripts", "00_packages.R"))

continuous_vars <- c(
  log10_years = "Elapsed years",
  log10_generations = "Elapsed generations"
)

categorical_vars <- c(
  disturbance = "Disturbance context",
  design = "Comparison design",
  trait_type = "Trait type",
  genphen = "Study response",
  data_scale = "Measurement scale"
)

theme_association <- function(base_size = 15) {
  ggplot2::theme_minimal(base_size = base_size) +
    ggplot2::theme(
      plot.title = ggplot2::element_text(face = "bold", size = base_size + 3),
      plot.subtitle = ggplot2::element_text(colour = "grey30"),
      panel.grid.minor = ggplot2::element_blank(),
      legend.position = "bottom",
      axis.text = ggplot2::element_text(colour = "grey15"),
      strip.text = ggplot2::element_text(face = "bold")
    )
}

cramers_v_corrected <- function(x, y) {
  tab <- table(x, y)
  n <- sum(tab)
  if (n < 2L || nrow(tab) < 2L || ncol(tab) < 2L) return(NA_real_)
  chi <- suppressWarnings(stats::chisq.test(tab, correct = FALSE))
  phi2 <- unname(chi$statistic) / n
  r <- nrow(tab)
  k <- ncol(tab)
  phi2_corrected <- max(0, phi2 - ((k - 1) * (r - 1)) / (n - 1))
  r_corrected <- r - ((r - 1)^2) / (n - 1)
  k_corrected <- k - ((k - 1)^2) / (n - 1)
  denom <- min(k_corrected - 1, r_corrected - 1)
  if (denom <= 0) NA_real_ else sqrt(phi2_corrected / denom)
}

eta_squared_test <- function(y, group) {
  keep <- stats::complete.cases(y, group)
  y <- y[keep]
  group <- droplevels(factor(group[keep]))
  if (length(y) < 3L || nlevels(group) < 2L) {
    return(tibble::tibble(estimate = NA_real_, statistic = NA_real_, p_value = NA_real_))
  }
  fit <- stats::lm(y ~ group)
  tab <- stats::anova(fit)
  tibble::tibble(
    estimate = tab$`Sum Sq`[1] / sum(tab$`Sum Sq`),
    statistic = tab$`F value`[1],
    p_value = tab$`Pr(>F)`[1]
  )
}

significance_mark <- function(p) {
  dplyr::case_when(
    is.na(p) ~ "",
    p < 0.001 ~ "***",
    p < 0.01 ~ "**",
    p < 0.05 ~ "*",
    TRUE ~ ""
  )
}
```

## Purpose and interpretation

This chapter checks whether the moderators overlap strongly enough to complicate their separate interpretation. Association is measured on the scale appropriate to each pair: Pearson's *r* for two continuous variables, bias-corrected Cramér's *V* for two categorical variables, and the correlation ratio η² for a continuous variable grouped by a categorical variable. Values closer to one indicate stronger association; they do not imply causation.

The tests below treat contrasts as rows and are therefore descriptive diagnostics rather than independent-observation causal tests. The study, system, and phylogenetic dependence handled by the meta-analytic models is not represented in these simple pairwise tests. Holm-adjusted *p*-values control multiplicity within this chapter.

```{r load-data}
dat_es <- readRDS(here::here("Rdata", "effect_sizes", "proceed_lnm_safe.rds"))

figure_dir <- here::here("Rdata", "figures", "moderator_associations")
table_dir <- here::here("Rdata", "tables", "moderator_associations")
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(table_dir, recursive = TRUE, showWarnings = FALSE)
```

## Continuous–continuous association

```{r continuous-analysis}
continuous_dat <- dat_es |>
  dplyr::select(dplyr::all_of(names(continuous_vars))) |>
  tidyr::drop_na()

pearson_test <- stats::cor.test(
  continuous_dat$log10_years,
  continuous_dat$log10_generations,
  method = "pearson"
)
spearman_test <- suppressWarnings(stats::cor.test(
  continuous_dat$log10_years,
  continuous_dat$log10_generations,
  method = "spearman",
  exact = FALSE
))

continuous_results <- tibble::tibble(
  method = c("Pearson", "Spearman sensitivity check"),
  estimate = c(unname(pearson_test$estimate), unname(spearman_test$estimate)),
  conf_low = c(pearson_test$conf.int[1], NA_real_),
  conf_high = c(pearson_test$conf.int[2], NA_real_),
  p_value = c(pearson_test$p.value, spearman_test$p.value),
  n = nrow(continuous_dat)
) |>
  dplyr::mutate(p_holm = stats::p.adjust(p_value, method = "holm"))

readr::write_csv(continuous_results, file.path(table_dir, "continuous_continuous.csv"))
knitr::kable(continuous_results, digits = 3,
             caption = "Continuous-moderator association and rank-based sensitivity check.")
```

```{r continuous-plot, fig.width=13, fig.height=9, fig.cap="Association between elapsed time expressed in years and generations. The annotation reports Pearson's correlation and its 95% confidence interval."}
continuous_label <- sprintf(
  "Pearson r = %.2f  |  95%% CI [%.2f, %.2f]  |  n = %s",
  continuous_results$estimate[1], continuous_results$conf_low[1],
  continuous_results$conf_high[1], scales::comma(continuous_results$n[1])
)

p_continuous <- ggplot2::ggplot(
  continuous_dat,
  ggplot2::aes(x = log10_years, y = log10_generations)
) +
  ggplot2::geom_point(colour = "#257180", alpha = 0.16, size = 1.7) +
  ggplot2::geom_smooth(method = "lm", formula = y ~ x, se = TRUE,
                       colour = "#CB6040", fill = "#F2E5BF", linewidth = 1.3) +
  ggplot2::labs(
    title = "Continuous moderators track a shared time scale",
    subtitle = continuous_label,
    x = expression(log[10]("elapsed years")),
    y = expression(log[10]("elapsed generations"))
  ) +
  theme_association(16)

ggplot2::ggsave(file.path(figure_dir, "continuous_continuous.png"), p_continuous,
                width = 13, height = 9, dpi = 320, bg = "white")
p_continuous
```

## Categorical–categorical associations

```{r categorical-analysis}
categorical_pairs <- utils::combn(names(categorical_vars), 2, simplify = FALSE)

categorical_results <- purrr::map_dfr(categorical_pairs, function(pair) {
  pair_dat <- dat_es |>
    dplyr::select(dplyr::all_of(pair)) |>
    tidyr::drop_na()
  tab <- table(pair_dat[[pair[1]]], pair_dat[[pair[2]]])
  chi <- suppressWarnings(stats::chisq.test(tab, correct = FALSE))
  tibble::tibble(
    variable_1 = pair[1], variable_2 = pair[2],
    label_1 = unname(categorical_vars[pair[1]]),
    label_2 = unname(categorical_vars[pair[2]]),
    cramers_v = cramers_v_corrected(pair_dat[[pair[1]]], pair_dat[[pair[2]]]),
    chi_square = unname(chi$statistic), df = unname(chi$parameter),
    p_value = chi$p.value, n = nrow(pair_dat),
    min_expected = min(chi$expected)
  )
}) |>
  dplyr::mutate(
    p_holm = stats::p.adjust(p_value, method = "holm"),
    stars = significance_mark(p_holm),
    tile_label = sprintf("%.2f%s", cramers_v, stars)
  )

readr::write_csv(categorical_results, file.path(table_dir, "categorical_categorical.csv"))
knitr::kable(
  categorical_results |>
    dplyr::select(label_1, label_2, cramers_v, chi_square, df, p_value, p_holm, n, min_expected),
  digits = 3,
  caption = "Pairwise categorical associations. Cramér's V is bias-corrected; p-values are from Pearson chi-square tests."
)
```

`r sum(categorical_results$min_expected < 5)` of `r nrow(categorical_results)` categorical comparisons contain at least one expected cell count below five. Their chi-square *p*-values should therefore be read cautiously; Cramér's *V*, which is the primary overlap diagnostic, remains the focus.

```{r categorical-plot, fig.width=14, fig.height=10, fig.cap="Bias-corrected Cramér's V among categorical moderators. Stars denote Holm-adjusted significance: * p < .05, ** p < .01, *** p < .001."}
cat_levels <- unname(categorical_vars)
p_categorical <- ggplot2::ggplot(
  categorical_results,
  ggplot2::aes(
    x = factor(label_2, levels = cat_levels),
    y = factor(label_1, levels = rev(cat_levels)),
    fill = cramers_v
  )
) +
  ggplot2::geom_tile(colour = "white", linewidth = 2) +
  ggplot2::geom_text(ggplot2::aes(label = tile_label, colour = cramers_v > 0.65),
                     fontface = "bold", size = 6) +
  ggplot2::scale_fill_gradientn(
    colours = c("#F7FBFF", "#9ECAE1", "#3182BD", "#08306B"),
    limits = c(0, 1), name = "Cramer's V"
  ) +
  ggplot2::scale_colour_manual(values = c(`FALSE` = "black", `TRUE` = "white"), guide = "none") +
  ggplot2::scale_x_discrete(drop = FALSE) +
  ggplot2::scale_y_discrete(drop = FALSE) +
  ggplot2::coord_fixed() +
  ggplot2::labs(
    title = "Overlap among categorical moderators",
    subtitle = "Larger values indicate stronger association; blank cells are redundant comparisons",
    x = NULL, y = NULL
  ) +
  theme_association(15) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 35, hjust = 1),
    panel.grid = ggplot2::element_blank(),
    legend.position = "right"
  )

ggplot2::ggsave(file.path(figure_dir, "categorical_categorical.png"), p_categorical,
                width = 14, height = 10, dpi = 320, bg = "white")
p_categorical
```

## Continuous–categorical associations

For each mixed pair, η² is the proportion of variation in the continuous moderator explained by the categorical grouping in a one-way model. Because η² is non-negative, the heatmap shows strength but not direction; group distributions should be consulted when direction matters.

```{r mixed-analysis}
mixed_grid <- tidyr::crossing(
  continuous = names(continuous_vars),
  categorical = names(categorical_vars)
)

mixed_results <- purrr::pmap_dfr(mixed_grid, function(continuous, categorical) {
  result <- eta_squared_test(dat_es[[continuous]], dat_es[[categorical]])
  result |>
    dplyr::mutate(
      continuous = continuous,
      categorical = categorical,
      continuous_label = unname(continuous_vars[continuous]),
      categorical_label = unname(categorical_vars[categorical]),
      n = sum(stats::complete.cases(dat_es[[continuous]], dat_es[[categorical]])),
      groups = nlevels(droplevels(factor(dat_es[[categorical]])))
    )
}) |>
  dplyr::mutate(
    p_holm = stats::p.adjust(p_value, method = "holm"),
    stars = significance_mark(p_holm),
    tile_label = sprintf("%.2f%s", estimate, stars)
  )

readr::write_csv(mixed_results, file.path(table_dir, "continuous_categorical.csv"))
knitr::kable(
  mixed_results |>
    dplyr::select(continuous_label, categorical_label, estimate, statistic,
                  p_value, p_holm, n, groups),
  digits = 3,
  col.names = c("Continuous moderator", "Categorical moderator", "Eta squared",
                "F", "p", "Holm p", "n", "Groups"),
  caption = "Mixed-variable associations. Eta squared is based on a one-way linear model."
)
```

```{r mixed-plot, fig.width=14, fig.height=7.5, fig.cap="Correlation ratio η² for every continuous–categorical moderator pair. Stars denote Holm-adjusted significance: * p < .05, ** p < .01, *** p < .001."}
p_mixed <- ggplot2::ggplot(
  mixed_results,
  ggplot2::aes(
    x = factor(categorical_label, levels = unname(categorical_vars)),
    y = factor(continuous_label, levels = rev(unname(continuous_vars))),
    fill = estimate
  )
) +
  ggplot2::geom_tile(colour = "white", linewidth = 2) +
  ggplot2::geom_text(ggplot2::aes(label = tile_label), fontface = "bold", size = 6.5) +
  ggplot2::scale_fill_gradientn(
    colours = c("#FFF7EC", "#FEC44F", "#D95F0E", "#7F2704"),
    limits = c(0, 1), name = expression(eta^2)
  ) +
  ggplot2::labs(
    title = "Categorical moderators explain differing amounts of time-scale variation",
    subtitle = "Each cell is the proportion of continuous-moderator variance explained by group membership",
    x = NULL, y = NULL
  ) +
  theme_association(16) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 30, hjust = 1),
    panel.grid = ggplot2::element_blank(),
    legend.position = "right"
  )

ggplot2::ggsave(file.path(figure_dir, "continuous_categorical.png"), p_mixed,
                width = 14, height = 7.5, dpi = 320, bg = "white")
p_mixed
```

## Reading the diagnostics

For screening purposes, effect-size magnitude is more informative than significance because the dataset contains thousands of contrasts. As rough descriptive guides, values near 0.1, 0.3, and 0.5 can be read as small, moderate, and large for *r* or Cramér's *V*. For η², values near 0.01, 0.06, and 0.14 are often described as small, moderate, and large. These are conventions rather than decision thresholds; substantive overlap, cell counts, and the dependency structure should guide model specification.
