---
title: "Dependency structure"
---

::: {.workflow}
::: {}
**Input**
Effect sizes and reconciled species names.
:::
::: {}
**Action**
Count links among effect sizes, references, systems, and species.
:::
::: {}
**Output**
Dependency tables used to define grouping terms.
:::
::: {}
**Next**
Check the phylogenetic matrix.
:::
:::

```{r setup}
source(here::here("Scripts", "00_packages.R"))
source(here::here("Scripts", "01_paths.R"))
source(here::here("Scripts", "05_phylogeny.R"))
```

## Input: build the dependency dataset

```{r load}
dat_es <- readRDS(here::here("Rdata", "effect_sizes", "proceed_lnm_safe.rds"))
name_map <- readRDS(here::here("Rdata", "phylogeny", "proceed_name_map.rds"))
dat_mapped <- apply_phylo_name_map(dat_es, name_map)

dat_re_diag <- dat_mapped |>
  dplyr::filter(
    !is.na(sp_ncbi_canonical),
    nzchar(as.character(sp_ncbi_canonical))
  ) |>
  dplyr::transmute(
    es_id_db = factor(es_id_db),
    ref_id   = factor(ref_id),
    sys_id   = factor(sys_id),
    sp_ncbi  = factor(sp_ncbi_canonical)
  ) |>
  tidyr::drop_na() |>
  droplevels()

analysis_counts <- tibble::tibble(
  Unit = c(
    "Contrasts",
    "Effect-size identifiers",
    "References",
    "Study systems",
    "Canonical species"
  ),
  N = c(
    nrow(dat_re_diag),
    dplyr::n_distinct(dat_re_diag$es_id_db),
    dplyr::n_distinct(dat_re_diag$ref_id),
    dplyr::n_distinct(dat_re_diag$sys_id),
    dplyr::n_distinct(dat_re_diag$sp_ncbi)
  )
)

knitr::kable(analysis_counts, caption = "Counts in the dependency dataset.")
```

“Canonical species” means unique species labels after taxonomic reconciliation.
The current data contain 261 resolved input names but 253 canonical species
because eight additional subspecies or duplicate labels collapse into five
shared species labels. The next chapter documents this reconciliation and
distinguishes the 253 current species from the 265 tips stored in the cached
phylogenetic matrix.

## Action: count pairwise links

```{r dependency-table}
summarise_links <- function(data, factor_a, factor_b) {
  links <- data |>
    dplyr::distinct(.data[[factor_a]], .data[[factor_b]]) |>
    dplyr::count(.data[[factor_a]], name = "n_b")

  tibble::tibble(
    `Group A` = factor_a,
    `Linked group B` = factor_b,
    `Levels of A` = dplyr::n_distinct(data[[factor_a]]),
    `Levels of B` = dplyr::n_distinct(data[[factor_b]]),
    `A levels linked to one B` = sum(links$n_b == 1),
    `A levels linked to one B (%)` = round(100 * mean(links$n_b == 1), 1),
    `Median B levels per A` = stats::median(links$n_b),
    `Maximum B levels per A` = max(links$n_b)
  )
}

group_vars <- c("es_id_db", "ref_id", "sys_id", "sp_ncbi")

dependency_table <- purrr::map_dfr(group_vars, function(a) {
  purrr::map_dfr(setdiff(group_vars, a), function(b) {
    summarise_links(dat_re_diag, a, b)
  })
})

knitr::kable(
  dependency_table,
  digits = 1,
  caption = "Counts and proportions for each directed pair of grouping variables."
)
```

## Study-system relationships

```{r system-table}
system_table <- dplyr::bind_rows(
  summarise_links(dat_re_diag, "sp_ncbi", "sys_id"),
  summarise_links(dat_re_diag, "sys_id", "sp_ncbi"),
  summarise_links(dat_re_diag, "ref_id", "sys_id"),
  summarise_links(dat_re_diag, "sys_id", "ref_id")
)

knitr::kable(
  system_table,
  digits = 1,
  caption = "Counts and proportions linking study systems with species and references."
)
```

```{r save, include=FALSE}
readr::write_csv(
  dependency_table,
  here::here("Rdata", "tables", "dependency_nesting_table.csv")
)
readr::write_csv(
  system_table,
  here::here("Rdata", "tables", "dependency_support_summary.csv")
)
readr::write_csv(
  analysis_counts,
  here::here("Rdata", "tables", "dependency_analysis_universe.csv")
)
```
