---
title: "Phylogenetic reconciliation"
---

```{r setup}
source(here::here("chapters", "Rscripts", "00_packages.R"))
source(here::here("chapters", "Rscripts", "01_paths.R"))
source(here::here("chapters", "Rscripts", "05_phylogeny.R"))
source(here::here("chapters", "Rscripts", "15_render_guard.R"))   # read-only: never rebuild the phylogeny here
```

## Why a phylogenetic random effect

Species that share evolutionary history tend to share trait values for reasons
unrelated to the moderators in these models. The phylogenetic random effect
`(1 | gr(sp_ncbi, cov = A))`, included in every main location–scale model,
accounts for this by letting species-level intercepts covary according to a
matrix `A` derived from a tree of the analysed species. This chapter documents
how that tree and `A` were built, and how species that could not be placed on
the tree are handled.

## Building the tree and the correlation matrix

Species names in PROCEED were matched to the Open Tree of Life synthetic tree
(Hinchliff et al. 2015) via the `rotl` package, retrieving the induced subtree
for the matched taxa. Because PROCEED spans plant, invertebrate, and vertebrate
species, name matching was not restricted to a single higher-level taxonomic
context.

The tree's tip labels were then reconciled against the original PROCEED species
names using a four-stage cascade implemented in `prepR4pcm`:

1. **Exact match**
2. **Normalized match** — case, whitespace, and underscore differences resolved
3. **Synonym resolution** — via GNVerifier (`authority = "gnverifier"`), checking
   the name against external taxonomic databases for accepted synonyms
4. **Fuzzy match** — approximate string matching for names still unresolved

Species that cannot be resolved at any stage are excluded from the
phylogenetic random effect (see [Unmatched species](#unmatched-species)
below), but are **not** excluded from the rest of the model.

Multifurcations in the retrieved tree were randomly resolved into bifurcations
with zero-length branches (`ape::multi2di`). Where branch lengths were not
available from Open Tree of Life, Grafen branch lengths (power ρ = 1) were
assigned (`ape::compute.brlen`). The resulting tree was converted to a
phylogenetic correlation matrix under a Brownian-motion model of trait
evolution (`prepR4pcm::pr_phylo_cor`, falling back to `ape::vcv.phylo` if
unavailable).

All of this logic lives in `R/05_phylogeny.R::build_or_read_phylogeny()`; this
chapter only *reads* the cached result — rebuilding it requires network access
to Open Tree of Life and GNVerifier and is not done automatically on render.

```{r recompute-toggle}
# Change to TRUE only to force a full network rebuild of the tree (slow,
# requires internet access). Normal renders read the cache below.
recompute_phylogeny <- FALSE
```

## Load cached phylogeny

```{r load-phylo}
tree_path      <- here::here("Rdata", "phylogeny", "proceed_tree.rds")
A_path         <- here::here("Rdata", "phylogeny", "proceed_A_matrix.rds")
name_map_path  <- here::here("Rdata", "phylogeny", "proceed_name_map.rds")
unmatched_path <- here::here("Rdata", "phylogeny", "unmatched_species.csv")

if (recompute_phylogeny) {
  dat_es <- readRDS(here::here("Rdata", "effect_sizes", "proceed_lnm_safe.rds"))
  phylo_out <- build_or_read_phylogeny(
    species_vec = unique(as.character(dat_es$sp_ncbi)),
    recompute   = TRUE
  )
} else {
  pace_need(A_path,        "phylogenetic correlation matrix")
  pace_need(name_map_path, "phylogenetic name map")
  phylo_out <- list(A = readRDS(A_path), name_map = readRDS(name_map_path))
}

A_full     <- phylo_out$A
name_map   <- phylo_out$name_map
tree_final <- tryCatch(readRDS(tree_path), error = function(e) NULL)
```

## Reconciliation summary

```{r reconciliation-summary}
n_total      <- nrow(name_map)
n_resolved   <- sum(!is.na(name_map$sp_ncbi_canonical) &
                       nzchar(name_map$sp_ncbi_canonical))
n_unresolved <- n_total - n_resolved
n_tips       <- if (!is.null(tree_final)) length(tree_final$tip.label) else nrow(A_full)

tibble::tibble(
  Metric = c(
    "Unique species names submitted",
    "Reconciled to a canonical taxon",
    "Unresolved (excluded from phylogenetic term)",
    "Unique tips in pruned tree / rows in A"
  ),
  N = c(n_total, n_resolved, n_unresolved, n_tips),
  `%` = round(100 * c(n_total, n_resolved, n_unresolved, n_tips) / n_total, 1)
)
```

## Unmatched species

These species names could not be reconciled to a tree tip at any stage of the
cascade (exact, normalized, synonym via GNVerifier, or fuzzy) and are
therefore excluded from the phylogenetic random effect. They are retained in
every other random effect.

```{r unmatched-table}
if (file.exists(unmatched_path)) {
  readr::read_csv(unmatched_path, show_col_types = FALSE)
} else {
  cat("No unmatched-species file found — all species were resolved.\n")
}
```

## Species retained but absent from the tree ("star" species)

A separate, smaller category of row loss can occur *inside* an individual
model: a species can be reconciled to a canonical name (and so appear in
`name_map`) but still be absent from the tree subset used to build `A` for a
particular moderator's data. `prepare_phylo_and_data()`, called once per model
in the location–scale model grid, assigns these species zero phylogenetic
correlation with every other species — a star-phylogeny contribution — rather
than dropping them. See the Dependency structure chapter for the per-contrast
accounting of this.

## Phylogenetic correlation matrix — summary

```{r a-matrix-summary}
off_diag <- A_full[row(A_full) != col(A_full)]
cat("A matrix dimensions:", nrow(A_full), "×", ncol(A_full), "\n")
cat("Off-diagonal correlation range: [",
    round(min(off_diag), 3), ", ", round(max(off_diag), 3), "]\n", sep = "")
```
