---
title: "Phylogenetic reconciliation"
---

```{r setup}
source(here::here("chapters", "Rscripts", "00_packages.R"))
source(here::here("chapters", "Rscripts", "01_paths.R"))
source(here::here("chapters", "Rscripts", "05_phylogeny.R"))
source(here::here("chapters", "Rscripts", "15_render_guard.R"))   # read-only: never rebuild the phylogeny here
```

## Why a phylogenetic random effect

Species that share evolutionary history tend to share trait values for reasons
unrelated to the moderators in these models. The phylogenetic random effect
`(1 | gr(sp_ncbi, cov = A))`, included in every main location–scale model,
accounts for this by letting species-level intercepts covary according to a
matrix `A` derived from a tree of the analysed species. This chapter documents
how that tree and `A` were built, and how species that could not be placed on
the tree are handled.

## Terms used in the species counts

- **Input name**: a distinct `sp_ncbi` label in the current effect-size data.
- **Resolved input name**: an input name successfully linked to a tree label.
- **Canonical species**: a unique reconciled label used as a model grouping
  level and as a row name when the phylogenetic matrix is subset.
- **Cached `A` tip**: any row retained in the saved phylogenetic correlation
  matrix. The cache can include species from an earlier data snapshot that are
  absent from the current effect-size data.

## Building the tree and the correlation matrix

Species names in PROCEED were matched to the Open Tree of Life synthetic tree
(Hinchliff et al. 2015) via the `rotl` package, retrieving the induced subtree
for the matched taxa. Because PROCEED spans plant, invertebrate, and vertebrate
species, name matching was not restricted to a single higher-level taxonomic
context.

The tree's tip labels were then reconciled against the original PROCEED species
names using a four-stage cascade implemented in `prepR4pcm`:

1. **Exact match**
2. **Normalized match** — case, whitespace, and underscore differences resolved
3. **Synonym resolution** — via GNVerifier (`authority = "gnverifier"`), checking
   the name against external taxonomic databases for accepted synonyms
4. **Fuzzy match** — approximate string matching for names still unresolved

Species that cannot be resolved at any stage receive no canonical label. Their
contrasts are excluded from the complete primary-model dataset before fitting
(see [Unmatched species](#unmatched-species) below).

Multifurcations in the retrieved tree were randomly resolved into bifurcations
with zero-length branches (`ape::multi2di`). Where branch lengths were not
available from Open Tree of Life, Grafen branch lengths (power ρ = 1) were
assigned (`ape::compute.brlen`). The resulting tree was converted to a
phylogenetic correlation matrix under a Brownian-motion model of trait
evolution (`prepR4pcm::pr_phylo_cor`, falling back to `ape::vcv.phylo` if
unavailable).

All of this logic lives in
`chapters/Rscripts/05_phylogeny.R::build_or_read_phylogeny()`; this
chapter only *reads* the cached result — rebuilding it requires network access
to Open Tree of Life and GNVerifier and is not done automatically on render.

## The `prepR4pcm` reconciliation functions {#prepr4pcm-source}

The cascade above is implemented in the `prepR4pcm` package
(`itchyshin/prepR4pcm`, v0.5.0.9000, commit `cca21b9`). Because those functions
are the black box behind `build_or_read_phylogeny()`, the exported functions
this chapter's pipeline actually calls — `pr_normalize_names()`,
`pr_get_tree()`, `reconcile_tree()`, `reconcile_apply()`, `reconcile_mapping()`,
and `pr_phylo_cor()` — are reproduced below for the record. The block is
**not evaluated** (`eval=FALSE`): rendering this chapter reads the cached
`.rds` reconciliation outputs (next section), it does not re-run the package.
Internal helpers (prefixed `.pr_*`, together with `pr_run_cascade()`,
`pr_align_tree()`, `pr_strip_authority()`, etc.) are not shown here; see the
[package source](https://github.com/itchyshin/prepR4pcm) for those.

```{r prepr4pcm-source, eval=FALSE}
# ── prepR4pcm exported functions used by build_or_read_phylogeny() ────────────
# Source: itchyshin/prepR4pcm v0.5.0.9000 (commit cca21b9). Reproduced for the
# record only; this chapter reads the cached .rds outputs instead of running it.

pr_normalize_names <- function (names, rank = c("species", "subspecies"), parser = c("internal",
    "gnparser"))
{
    rank <- match.arg(rank)
    parser <- match.arg(parser)
    if (parser == "gnparser") {
        return(.pr_normalize_gnparser(names, rank))
    }
    original <- names
    names <- as.character(names)
    is_na <- is.na(names)
    names[is_na] <- ""
    names <- trimws(names)
    names <- gsub("_", " ", names, fixed = TRUE)
    names <- gsub("\\s+", " ", names)
    names <- gsub("\\s+ott\\d+$", "", names, perl = TRUE)
    names <- pr_strip_authority(names)
    names <- sub("\\s*\\([^()]*\\)\\s*$", "", names, perl = TRUE)
    names <- gsub("\\s*×\\s*", " x ", names)
    names <- gsub("^x\\s+", "x ", names)
    names <- gsub("\\s+x\\s+", " x ", names)
    names <- gsub("\\bsubsp\\.?\\s+", "subsp. ", names, perl = TRUE)
    names <- gsub("\\bssp\\.?\\s+", "subsp. ", names, perl = TRUE)
    names <- gsub("\\bvar\\.?\\s+", "var. ", names, perl = TRUE)
    names <- gsub("\\bf\\.?\\s+", "f. ", names, perl = TRUE)
    if (rank == "species") {
        names <- pr_strip_infraspecific(names)
    }
    names <- pr_standardise_case(names)
    names <- trimws(names)
    names[is_na] <- NA_character_
    changed <- !is_na & (original != names)
    log <- tibble(original = original, normalised = names, changed = changed)
    attr(names, "normalisation_log") <- log
    names
}

pr_get_tree <- function (x, source = c("rotl", "rtrees", "clootl", "fishtree",
    "datelife", "auto"), species_col = NULL, taxon = NULL, n_tree = 1L,
    cache = FALSE, tnrs = c("auto", "always", "never"), min_match = 0.8,
    check_ultrametric = TRUE, resolve_polytomies = FALSE, branch_lengths = NULL,
    ...)
{
    source <- match.arg(source)
    tnrs <- match.arg(tnrs)
    if (!is.null(branch_lengths)) {
        branch_lengths <- match.arg(branch_lengths, choices = c("grafen",
            "compute.brlen", "unit"))
    }
    n_tree <- .pr_validate_positive_integer(n_tree, "n_tree")
    if (!is.numeric(min_match) || length(min_match) != 1L ||
        is.na(min_match) || !is.finite(min_match) || min_match <
        0 || min_match > 1) {
        cli::cli_abort(c("{.arg min_match} must be a length-1 numeric in [0, 1].",
            i = "Got: {.val {min_match}}."))
    }
    species <- .pr_extract_species_for_tree(x, species_col)
    if (length(species) == 0) {
        cli::cli_abort(c("No species names available to query the backend.",
            i = "If you passed a {.cls reconciliation} object, ensure {.code mapping$name_y} contains resolved names."))
    }
    if (source == "auto") {
        return(.pr_get_tree_auto(species, taxon = taxon, n_tree = n_tree,
            cache = cache, tnrs = tnrs, min_match = min_match,
            ...))
    }
    resolved <- .pr_resolve_query(species, source, tnrs)
    species_query <- resolved$query
    if (isTRUE(cache)) {
        key <- .pr_tree_cache_key(resolved$original, source = source,
            n_tree = n_tree, taxon = taxon, tnrs = tnrs, query = species_query,
            ...)
        cached <- .pr_tree_cache_get(key, source)
        if (!is.null(cached)) {
            return(cached)
        }
    }
    result <- switch(source, rotl = .pr_get_tree_rotl(species_query,
        n_tree = n_tree, ...), rtrees = .pr_get_tree_rtrees(species_query,
        n_tree = n_tree, taxon = taxon, ...), clootl = .pr_get_tree_clootl(species_query,
        n_tree = n_tree, ...), fishtree = .pr_get_tree_fishtree(species_query,
        n_tree = n_tree, ...), datelife = .pr_get_tree_datelife(species_query,
        n_tree = n_tree, ...))
    in_query <- result$in_query
    if (!is.logical(in_query) || length(in_query) != length(resolved$original)) {
        cli::cli_abort(c("Internal: backend wrapper returned malformed {.code in_query}.",
            i = "Expected {.cls logical} of length {length(resolved$original)}; got {.cls {class(in_query)[1]}} of length {length(in_query)}."))
    }
    matched <- resolved$original[in_query]
    unmatched <- resolved$original[!in_query]
    is_multi <- inherits(result$tree, "multiPhylo")
    n_returned <- if (is_multi)
        length(result$tree)
    else 1L
    tip_set_consistent <- TRUE
    dropped_per_tree <- NULL
    if (is_multi && length(result$tree) > 1L) {
        tip_sets <- lapply(result$tree, function(t) sort(t$tip.label))
        tip_set_consistent <- all(vapply(tip_sets[-1L], function(s) identical(s,
            tip_sets[[1L]]), logical(1L)))
        if (!tip_set_consistent) {
            union_tips <- Reduce(union, tip_sets)
            dropped_per_tree <- lapply(tip_sets, function(s) setdiff(union_tips,
                s))
        }
    }
    stopifnot(`matched must be a subset of unique input` = all(matched %in%
        resolved$original), `unmatched must be a subset of unique input` = all(unmatched %in%
        resolved$original), `matched + unmatched must cover unique input` = length(matched) +
        length(unmatched) == length(resolved$original), `matched and unmatched must be disjoint` = length(intersect(matched,
        unmatched)) == 0L)
    wrapper_meta <- if (is.list(result$backend_meta)) {
        result$backend_meta
    }
    else {
        list()
    }
    if (identical(source, "rtrees") && is.data.frame(wrapper_meta$placement) &&
        nrow(wrapper_meta$placement) == length(resolved$original)) {
        wrapper_meta$placement$input_name <- resolved$original
    }
    backend_meta <- utils::modifyList(wrapper_meta, list(n_queried = length(resolved$original),
        n_requested = as.integer(n_tree), n_returned = n_returned,
        n_matched = length(matched), tnrs_replacements = resolved$tnrs_replacements,
        tip_set_consistent = tip_set_consistent, dropped_per_tree = dropped_per_tree))
    backend_meta <- .pr_ensure_tree_provenance(result$tree, backend_meta,
        source)
    result$tree <- .pr_post_process_tree(result$tree, resolve_polytomies = resolve_polytomies,
        branch_lengths = branch_lengths)
    mapping <- .pr_build_tree_mapping(input_name = resolved$original,
        normalized_name = resolved$normalised, query_name = resolved$query,
        in_tree = in_query, tree = result$tree, backend_meta = backend_meta,
        tnrs_audit = resolved$tnrs_audit)
    out <- list(tree = result$tree, matched = matched, unmatched = unmatched,
        mapping = mapping, source = source, backend_meta = backend_meta)
    class(out) <- "pr_tree_result"
    if (isTRUE(cache)) {
        .pr_tree_cache_put(key, source, out)
    }
    if (isTRUE(check_ultrametric)) {
        .pr_check_tree_ultrametric(out$tree, source)
    }
    out
}

reconcile_tree <- function (x, tree, x_species = NULL, authority = "col", rank = c("species",
    "subspecies"), overrides = NULL, db_version = NULL, fuzzy = FALSE,
    fuzzy_threshold = 0.9, flag_threshold = 0.95, resolve = c("flag",
        "first"), quiet = FALSE, x_label = NULL)
{
    x_source <- x_label %||% deparse(substitute(x))
    rank <- match.arg(rank)
    resolve <- match.arg(resolve)
    if (!is.data.frame(x))
        abort("`x` must be a data frame.", call = caller_env())
    authority <- pr_validate_authority(authority)
    if (is.null(x_species))
        x_species <- pr_detect_species_column(x, "x_species")
    if (!x_species %in% names(x)) {
        abort(paste0("Column '", x_species, "' not found in `x`."),
            call = caller_env())
    }
    tree_obj <- pr_load_tree(tree)
    tips <- tree_obj$tip.label
    if (nrow(x) == 0)
        abort("`x` has 0 rows.", call = caller_env())
    if (is.factor(x[[x_species]])) {
        cli_alert_warning("Converting factor column '{x_species}' in `x` to character.")
        x[[x_species]] <- as.character(x[[x_species]])
    }
    names_x <- as.character(x[[x_species]])
    if (all(is.na(names_x))) {
        abort("All species names in `x` are NA.", call = caller_env())
    }
    tree_source <- if (is.character(tree)) {
        basename(tree)
    }
    else {
        sprintf("phylo (%d tips)", length(tips))
    }
    overrides_df <- pr_load_overrides(overrides)
    if (!quiet) {
        cli_alert_info("Reconciling {length(unique(names_x))} data names vs {length(tips)} tree tips")
    }
    mapping <- pr_run_cascade(names_x = names_x, names_y = tips,
        authority = authority, db_version = db_version, rank = rank,
        overrides = overrides_df, fuzzy = fuzzy, fuzzy_threshold = fuzzy_threshold,
        flag_threshold = flag_threshold, resolve = resolve, quiet = quiet)
    meta <- list(call = match.call(), type = "data_tree", timestamp = Sys.time(),
        authority = authority %||% "none", db_version = db_version %||%
            "latest", fuzzy = fuzzy, fuzzy_threshold = if (fuzzy) fuzzy_threshold else NA_real_,
        fuzzy_method = if (fuzzy) "component_levenshtein" else NA_character_,
        resolve = resolve, prepR4pcm_version = as.character(utils::packageVersion("prepR4pcm")),
        x_source = x_source, y_source = tree_source, rank = rank)
    result <- new_reconciliation(mapping = mapping, meta = meta)
    if (!quiet && nrow(result$unused_overrides) > 0) {
        pr_warn_unused_overrides(result$unused_overrides)
    }
    if (!quiet) {
        n_matched <- sum(mapping$in_x & mapping$in_y, na.rm = TRUE)
        n_total <- result$counts$n_x
        cli_alert_success("Matched {n_matched}/{n_total} data names to tree tips")
    }
    result
}

reconcile_apply <- function (reconciliation, data = NULL, tree = NULL, species_col = NULL,
    drop_unresolved = FALSE)
{
    validate_reconciliation(reconciliation)
    mapping <- reconciliation$mapping
    matched <- mapping[mapping$in_x & mapping$in_y, ]
    result <- list(data = NULL, tree = NULL)
    if (!is.null(data)) {
        if (!is.data.frame(data)) {
            abort("`data` must be a data frame.", call = caller_env())
        }
        if (is.null(species_col)) {
            species_col <- pr_detect_species_column(data, "species_col")
        }
        if (!species_col %in% names(data)) {
            cli::cli_abort(c("Column {.val {species_col}} not found in {.arg data}.",
                i = "Available columns: {.val {names(data)}}."))
        }
        data_names <- as.character(data[[species_col]])
        if (drop_unresolved) {
            keep <- data_names %in% matched$name_x
            data <- data[keep, , drop = FALSE]
            n_dropped <- sum(!keep)
            if (n_dropped > 0) {
                cli_alert_warning("Dropped {n_dropped} rows with unresolved species from data")
            }
        }
        result$data <- data
    }
    if (!is.null(tree)) {
        tree <- pr_load_tree(tree)
        tree <- pr_align_tree(tree, mapping, drop_unresolved = drop_unresolved)
        if (drop_unresolved) {
            n_tips <- length(tree$tip.label)
            cli_alert_info("Tree has {n_tips} tips after alignment")
        }
        result$tree <- tree
    }
    result
}

reconcile_mapping <- function (reconciliation, include_unused_overrides = FALSE)
{
    validate_reconciliation(reconciliation)
    if (!isTRUE(include_unused_overrides)) {
        return(reconciliation$mapping)
    }
    unused <- reconciliation$unused_overrides
    if (is.null(unused) || nrow(unused) == 0) {
        return(reconciliation$mapping)
    }
    unused_rows <- tibble(name_x = unused$name_x, name_y = unused$name_y,
        name_resolved = NA_character_, match_type = "override_unused",
        match_score = NA_real_, match_source = "user_override",
        in_x = FALSE, in_y = FALSE, notes = unused$reason)
    rbind(reconciliation$mapping, unused_rows)
}

pr_phylo_cor <- function (x, corr = TRUE, ...)
{
    tree <- if (inherits(x, "pr_tree_result"))
        x$tree
    else x
    if (inherits(tree, "multiPhylo")) {
        return(lapply(tree, pr_phylo_cor, corr = corr, ...))
    }
    if (!inherits(tree, "phylo")) {
        cli::cli_abort(c("{.arg x} must be a {.cls phylo}, {.cls multiPhylo}, or {.cls pr_tree_result}.",
            i = "Got: {.cls {class(x)[1]}}."))
    }
    if (is.null(tree$edge.length)) {
        cli::cli_abort(c("Tree has no branch lengths --- {.fn pr_phylo_cor} cannot compute a correlation matrix.",
            i = "If your tree came from {.code source = \"rotl\"}, run {.fn pr_get_tree} again with {.code branch_lengths = \"grafen\"} to assign Grafen's method-based lengths.",
            `>` = "Or transform manually: {.code tree <- ape::compute.brlen(tree, method = \"Grafen\")}."))
    }
    ape::vcv(tree, corr = corr, ...)
}
```

```{r recompute-toggle}
# Change to TRUE only to force a full network rebuild of the tree (slow,
# requires internet access). Normal renders read the cache below.
recompute_phylogeny <- FALSE
```

## Load cached phylogeny

```{r load-phylo}
tree_path      <- here::here("Rdata", "phylogeny", "proceed_tree.rds")
A_path         <- here::here("Rdata", "phylogeny", "proceed_A_matrix.rds")
name_map_path  <- here::here("Rdata", "phylogeny", "proceed_name_map.rds")
unmatched_path <- here::here("Rdata", "phylogeny", "unmatched_species.csv")

if (recompute_phylogeny) {
  dat_es <- readRDS(here::here("Rdata", "effect_sizes", "proceed_lnm_safe.rds"))
  phylo_out <- build_or_read_phylogeny(
    species_vec = unique(as.character(dat_es$sp_ncbi)),
    recompute   = TRUE
  )
} else {
  pace_need(A_path,        "phylogenetic correlation matrix")
  pace_need(name_map_path, "phylogenetic name map")
  phylo_out <- list(A = readRDS(A_path), name_map = readRDS(name_map_path))
}

A_full     <- phylo_out$A
name_map   <- phylo_out$name_map
tree_final <- tryCatch(readRDS(tree_path), error = function(e) NULL)

# Use the current effect-size dataset for reported analysis counts. The cached
# scaffold can contain tips from an earlier, larger data snapshot.
dat_es <- readRDS(here::here("Rdata", "effect_sizes", "proceed_lnm_safe.rds"))
dat_mapped <- apply_phylo_name_map(dat_es, name_map)
```

## Reconciliation counts

```{r reconciliation-summary}
raw_species <- unique(as.character(dat_mapped$sp_ncbi))
raw_species <- raw_species[!is.na(raw_species) & nzchar(trimws(raw_species))]
current_map <- name_map |>
  dplyr::filter(sp_ncbi_original %in% raw_species)
current_canonical <- unique(as.character(dat_mapped$sp_ncbi_canonical))
current_canonical <- current_canonical[
  !is.na(current_canonical) & nzchar(current_canonical)
]

n_current_names <- length(raw_species)
n_current_resolved_names <- sum(
  !is.na(current_map$sp_ncbi_canonical) & nzchar(current_map$sp_ncbi_canonical)
)
n_current_unresolved_names <- n_current_names - n_current_resolved_names
n_current_tips <- length(intersect(current_canonical, rownames(A_full)))
n_cached_tips <- nrow(A_full)
n_names_combined <- n_current_resolved_names - length(current_canonical)

current_counts <- tibble::tibble(
  Metric = c(
    "Distinct input names in current effect-size dataset",
    "Input names successfully resolved",
    "Input names unresolved",
    "Additional resolved names combined under shared canonical labels",
    "Unique canonical species used by current models",
    "Contrasts used by current models",
    "Contrasts excluded because their species name was unresolved"
  ),
  N = c(
    n_current_names, n_current_resolved_names, n_current_unresolved_names,
    n_names_combined, n_current_tips,
    sum(!is.na(dat_mapped$sp_ncbi_canonical)),
    sum(is.na(dat_mapped$sp_ncbi_canonical))
  )
)

scaffold_counts <- tibble::tibble(
  Metric = c(
    "All tips stored in cached A matrix",
    "Cached A tips used by current dataset",
    "Cached A tips not used by current dataset"
  ),
  N = c(
    n_cached_tips,
    n_current_tips,
    length(setdiff(rownames(A_full), current_canonical))
  )
)

knitr::kable(
  current_counts,
  caption = "Current-data reconciliation: 267 input labels become 253 model species."
)

knitr::kable(
  scaffold_counts,
  caption = "Cached phylogenetic scaffold: only 253 of its 265 tips are used by the current data."
)
```

The count of 253 therefore describes the current model dataset. The count of
265 describes the reusable cached matrix, not the current sample. The 12
unused cached tips are removed when `A` is subset for a model.

### Input names combined under canonical species

```{r canonical-collapses}
collapse_table <- current_map |>
  dplyr::filter(
    !is.na(sp_ncbi_canonical),
    nzchar(sp_ncbi_canonical)
  ) |>
  dplyr::group_by(sp_ncbi_canonical) |>
  dplyr::summarise(
    `Input names` = paste(sort(sp_ncbi_original), collapse = "; "),
    `Number of input names` = dplyr::n(),
    .groups = "drop"
  ) |>
  dplyr::filter(`Number of input names` > 1) |>
  dplyr::rename(`Canonical species` = sp_ncbi_canonical)

knitr::kable(
  collapse_table,
  caption = "Resolved input labels that share a canonical species label."
)
```

## Unmatched species

These current species names could not be reconciled to a tree tip at any stage
of the cached cascade (exact, normalized, synonym via GNVerifier, or fuzzy).
Because primary models require the phylogenetic term, their contrasts are
excluded from the complete fitted row set before model fitting.

```{r unmatched-table}
if (file.exists(unmatched_path)) {
  readr::read_csv(unmatched_path, show_col_types = FALSE) |>
    dplyr::filter(sp_ncbi_original %in% raw_species)
} else {
  cat("No unmatched-species file found — all species were resolved.\n")
}
```

## Species retained but absent from the tree ("star" species)

A separate, smaller category of row loss can occur *inside* an individual
model: a species can be reconciled to a canonical name (and so appear in
`name_map`) but still be absent from the tree subset used to build `A` for a
particular moderator's data. `prepare_phylo_and_data()`, called once per model
in the location–scale model grid, assigns these species zero phylogenetic
correlation with every other species — a star-phylogeny contribution — rather
than dropping them. See the Dependency structure chapter for the per-contrast
accounting of this. This applies only to a non-missing canonical name absent
from `A`; it does not apply to an unresolved original name. The current data
contain `r length(setdiff(current_canonical, rownames(A_full)))` star species.

## Phylogenetic correlation matrix — summary

```{r a-matrix-summary}
off_diag <- A_full[row(A_full) != col(A_full)]
cat("A matrix dimensions:", nrow(A_full), "×", ncol(A_full), "\n")
cat("Off-diagonal correlation range: [",
    round(min(off_diag), 3), ", ", round(max(off_diag), 3), "]\n", sep = "")
```
