---
title: "Sensitivity analyses"
---

::: {.workflow}
::: {}
**Input**
Primary and sensitivity summary tables.
:::
::: {}
**Action**
Apply the recorded comparison rule.
:::
::: {}
**Output**
Status counts and paired coefficient tables.
:::
::: {}
**Next**
Create final figures.
:::
:::

```{r setup, include=FALSE}
source(here::here("Scripts", "00_packages.R"))
source(here::here("Scripts", "19_sensitivity_tables.R"))
```

## Comparison setup

The seven moderator models were repeated under three prespecified alternatives:

| Analysis | Recorded change | Fixed comparison |
|---|---|---|
| Minimum sample size | Keep only contrasts with $n_1 + n_2 \geq 40$ | Matching primary model |
| No phylogeny | Remove the phylogenetic random effect | Matching primary model |
| Study-system effect | Add a `sys_id` random intercept | Matching primary model |

Each analysis changes **one feature at a time**. The response, moderator
definitions, priors, study-level (`ref_id`) effect, and MCMC settings otherwise
match the corresponding comparison model. The no-phylogeny analysis necessarily omits the
phylogenetic effect; the minimum-sample-size analysis necessarily uses fewer
contrasts.

## At-a-glance results

The table below is generated from the latest synced model summaries. “Models
available” should equal seven for a complete analysis. “Changed coefficients”
excludes intercepts and counts a coefficient when either:

1. its posterior median changes sign; or
2. its 95% credible interval changes from including zero to excluding zero, or
   vice versa.

This rule creates the displayed `DIFFERS` and `consistent` labels. The paired
estimates are also printed in the appendices.

```{r sensitivity-summary}
sens_variant_summary_table()
```

## Output columns

Each appendix begins with the exact model change and its convergence
diagnostics, then places every sensitivity coefficient beside the corresponding
primary coefficient:

- [Minimum total sample size ($n_1 + n_2 \geq 40$)](appendix/sensitivity_min_n.qmd)
- [No phylogenetic random effect](appendix/sensitivity_no_phylogeny.qmd)
- [Study-system (`sys_id`) random effect](appendix/sensitivity_sys_id.qmd)

The tables contain two submodels:

- **Location (mean lnM):** effects on the expected magnitude of divergence.
- **Scale (log-SD):** effects on between-observation variability, expressed on
  the log standard-deviation scale.

These are raw treatment-coded coefficients, not marginal means. For categorical
moderators, the intercept is the reference-level estimate and every other
coefficient is a difference from that reference. For continuous moderators,
the non-intercept coefficient is the slope. Intercepts are shown for
completeness but excluded from the summary change count.

Rows marked **DIFFERS** meet one of the two screening criteria above. Rows
marked *consistent* do not; the label does not mean the estimates are
numerically identical.
