---
title: "Sensitivity analyses"
---

```{r setup, include=FALSE}
source(here::here("chapters", "Rscripts", "00_packages.R"))
source(here::here("chapters", "Rscripts", "19_sensitivity_tables.R"))
```

## Model status and overall results

The three prespecified sensitivity analyses were run for each of the seven
moderators included in the primary model set. The table below is generated from
the synced model summaries rather than entered by hand. A changed coefficient
is one whose sign or 95% credible-interval conclusion differs from its matched
primary-model coefficient; intercepts are excluded from this count.

```{r sensitivity-summary, echo=FALSE}
sens_variant_summary_table()
```

All 21 expected sensitivity models have cached results. Removing the
phylogenetic random effect produced the fewest changes relative to the primary
fits. The minimum-sample-size and study-system analyses produced more changes,
and their divergent transitions mean those differences should be interpreted
alongside the model-specific diagnostics rather than as definitive reversals.

## Detailed results

The appendices report every location and scale coefficient beside the matched
primary estimate, together with convergence diagnostics:

- [Minimum total sample size (N >= 40)](appendix/sensitivity_min_n.qmd)
- [No phylogenetic random effect](appendix/sensitivity_no_phylogeny.qmd)
- [Study-system (`sys_id`) random effect](appendix/sensitivity_sys_id.qmd)

Rows marked **DIFFERS** identify coefficients whose sign or credible-interval
conclusion changes. The detailed tables use raw treatment-coded
coefficients so that each sensitivity estimate is compared like-for-like with
the corresponding primary fit.
