---
title: "Small-study effects diagnostic"
---

::: {.workflow}
::: {}
**Input**
SAFE lnM estimates, group sample sizes, and the phylogenetic covariance matrix.
:::
::: {}
**Action**
Fit sample-size-based Egger-style multilevel meta-regressions.
:::
::: {}
**Output**
Small-study slope, infinite-sample intercept, and diagnostic plot.
:::
::: {}
**Next**
Check whether conclusions are robust to alternative analysis specifications.
:::
:::

```{r setup, include=FALSE}
source(here::here("chapters", "Rscripts", "00_packages.R"))
source(here::here("chapters", "Rscripts", "20_small_study_effects.R"))
library(metafor)

result_path <- here::here("Rdata", "tables", "small_study_effects.csv")
model_path <- here::here("Rdata", "models", "small_study_effects.rds")
figure_path <- here::here("Rdata", "figures", "publication", "small_study_effects.png")

if (!file.exists(result_path) || !file.exists(model_path) || !file.exists(figure_path)) {
  stop(
    "Small-study-effects outputs are missing. Run ",
    "run_small_study_analysis() from chapters/Rscripts/20_small_study_effects.R.",
    call. = FALSE
  )
}

small_study_results <- readr::read_csv(result_path, show_col_types = FALSE)
small_study_fit <- readRDS(model_path)
```

## Why sample size, rather than sampling error?

We assess whether estimated effect magnitudes vary systematically with study
size. Such a pattern is a **small-study effect**: it can be consistent with
publication bias, but it can also arise from genuine heterogeneity or design
differences. It is therefore a diagnostic, not a direct test of the publication
process.

The usual Egger regression uses an effect's sampling standard error as the
predictor. For lnM, however, the effect and its estimated standard error share
underlying variance components. Their structural dependence can create
artefactual asymmetry. Following the
[lnM worked example](https://itchyshin.github.io/meta-analysis_of_magnitude/)
and Nakagawa et al. (2022), we instead use a predictor calculated only from the
two group sample sizes.

For each contrast, the harmonic effective sample size is

$$
n_0 = \frac{n_1 n_2}{n_1 + n_2}.
$$

We fit two related models:

1. $1/\sqrt{n_0}$ is an SE analogue; its slope is the primary diagnostic for
   small-study effects.
2. $1/n_0$ is a variance analogue; its intercept at $1/n_0=0$ is an
   infinite-sample, bias-adjusted estimate. We treat this as a conditional
   sensitivity estimate rather than an automatic replacement for the primary
   meta-analytic mean.

### Constructing the predictors

The two predictors are calculated directly from the original group sample
sizes. Neither uses the observed lnM estimate or its sampling variance.

```{r predictor-code, eval=FALSE}
diagnostic_data <- readRDS(
  here::here("Rdata", "effect_sizes", "proceed_lnm_safe.rds")
) |>
  dplyr::filter(
    is.finite(n1), n1 > 0,
    is.finite(n2), n2 > 0
  ) |>
  dplyr::mutate(
    n0   = (n1 * n2) / (n1 + n2),
    n_se = 1 / sqrt(n0),
    n_v  = 1 / n0
  )
```

## Model structure

Both models use SAFE lnM as the response and retain the analysis's known
sampling variances. Random intercepts account for clustering by reference,
phylogenetically correlated species, and effect-size record. Thus the
diagnostic changes the fixed predictor while preserving the main sources of
non-independence.

```{r model-code, eval=FALSE}
diagnostic_data <- prepare_small_study_data()

fit_se <- metafor::rma.mv(
  yi = yi_lnM_safe,
  V = vi_lnM_safe,
  mods = ~ n_se,
  random = list(
    ~ 1 | ref_id,
    ~ 1 | sp_ncbi,
    ~ 1 | es_id_model
  ),
  R = list(sp_ncbi = diagnostic_data$A),
  Rscale = "cor",
  method = "REML",
  test = "t",
  sparse = TRUE,
  data = diagnostic_data$data
)

fit_v <- metafor::rma.mv(
  yi = yi_lnM_safe,
  V = vi_lnM_safe,
  mods = ~ n_v,
  random = list(
    ~ 1 | ref_id,
    ~ 1 | sp_ncbi,
    ~ 1 | es_id_model
  ),
  R = list(sp_ncbi = diagnostic_data$A),
  Rscale = "cor",
  method = "REML",
  test = "t",
  sparse = TRUE,
  data = diagnostic_data$data
)
```

The diagnostic includes **`r small_study_fit$n` contrasts** with complete,
positive group sample sizes and resolved species names.

## Results

### Model summaries

The following are the complete `summary()` outputs from the two cached model
objects fitted on Totoro.

```{r model-summary-se}
summary(small_study_fit$fit_se)
```

```{r model-summary-v}
summary(small_study_fit$fit_v)
```

### Coefficient table

```{r results-table}
display_results <- small_study_results |>
  dplyr::mutate(
    term = dplyr::recode(
      term,
      intrcpt = "Intercept",
      n_se = "1 / sqrt(n0) slope",
      n_v = "1 / n0 slope"
    ),
    across(c(estimate, std_error, ci_lower, ci_upper), ~ round(.x, 3)),
    p_value = format.pval(p_value, digits = 3, eps = 0.001)
  ) |>
  dplyr::select(Model = model, Term = term, Estimate = estimate,
                `Standard error` = std_error, `95% CI lower` = ci_lower,
                `95% CI upper` = ci_upper, `p-value` = p_value)

knitr::kable(
  display_results,
  caption = paste0(
    "Sample-size-based multilevel meta-regressions. The primary diagnostic ",
    "is the 1 / sqrt(n0) slope; the 1 / n0 intercept estimates the effect at ",
    "infinite sample size."
  )
)
```

```{r interpretation, results='asis'}
slope <- small_study_results |>
  dplyr::filter(model == "Small-study slope: 1 / sqrt(n0)", term == "n_se")

direction <- if (slope$estimate > 0) "larger" else "smaller"
evidence <- if (slope$ci_lower > 0 || slope$ci_upper < 0) {
  "The 95% confidence interval excludes zero, providing evidence of a small-study effect"
} else {
  "The 95% confidence interval includes zero, so this analysis does not provide clear evidence of a small-study effect"
}

cat(sprintf(
  "%s (slope = %.3f, 95%% CI [%.3f, %.3f]). The fitted direction associates smaller effective samples with %s lnM estimates. This pattern should not, by itself, be labelled publication bias.\n",
  evidence, slope$estimate, slope$ci_lower, slope$ci_upper, direction
))
```

## Diagnostic plot

The two panels are constructed from the cached models and joined with
`patchwork`. Bubble area represents sampling precision, $1/\mathrm{SE}$.

```{r figure-code, eval=FALSE}
diagnostic_data <- prepare_small_study_data()

p_se <- build_small_study_figure(
  small_study_fit$fit_se,
  diagnostic_data$data,
  predictor = "n_se",
  bubble_fill = "#DC143C",
  line_colour = "#DC143C"
)

p_v <- build_small_study_figure(
  small_study_fit$fit_v,
  diagnostic_data$data,
  predictor = "n_v",
  bubble_fill = "#FF8C00",
  line_colour = "#8B4500"
) +
  ggplot2::theme(legend.position = "none")

p <- ((p_se + p_v) / patchwork::guide_area()) +
  patchwork::plot_layout(heights = c(1, 0.10), guides = "collect") +
  patchwork::plot_annotation(tag_levels = "A", tag_suffix = ")") &
  ggplot2::theme(
    legend.position = "bottom",
    legend.justification = "center",
    plot.tag = ggplot2::element_text(face = "bold", size = 14),
    plot.tag.position = c(0.01, 0.99)
  )

ggplot2::ggsave(
  here::here("Rdata", "figures", "publication", "small_study_effects.png"),
  p, width = 13, height = 6, dpi = 300, bg = "white"
)
```

```{r figure}
knitr::include_graphics(figure_path)
```

::: {.callout-note}
The infinite-sample intercept is an extrapolation to $1/n_0=0$, outside the
observed finite-sample data. It is most useful as a sensitivity analysis when
the slope diagnostic indicates small-study effects and the model is otherwise
well specified.
:::

## Reproduce the cached outputs

Normal book renders read the saved model, table, and figure. To refit the
diagnostic deliberately, run:

```{r reproduce, eval=FALSE}
source(here::here("chapters", "Rscripts", "00_packages.R"))
source(here::here("chapters", "Rscripts", "20_small_study_effects.R"))
run_small_study_analysis()
```
