---
title: "Reproduce this analysis"
---

::: {.workflow}
::: {}
**Input**
The complete project folder.
:::
::: {}
**Action**
Install software, choose a route, and render.
:::
::: {}
**Output**
The web book and an environment record.
:::
::: {}
**Next**
Check every item in the final checklist.
:::
:::

## Before you start {#sec-reproducibility}

This chapter gives the setup steps in order. Run all commands from the project
folder. Use the **quick route** to rebuild the web book from saved results. Use
the **full route** to calculate the effect sizes and fit the models again.

::: {.callout-important title="Use the project folder"}
Open `Magnitude_pace_modern_life.Rproj` before running R code. Do not change the
folder names. The code uses `here::here()` to find files from the project root.
:::

## Step 1 — Get the project

Download the repository from GitHub or clone it with Git:

```bash
git clone ANONYMISED-REPOSITORY-URL
cd Magnitude_pace_modern_life
```

Keep the `data/`, `Rdata/`, `chapters/`, and `_quarto.yml` paths unchanged.

## Step 2 — Install R packages

The project checks one package list in `Scripts/00_packages.R`. The
list includes every non-base R package named in the book and analysis scripts,
including `ggbeeswarm` and `patchwork`.

```r
cran_packages <- c(
  "ape", "bayesplot", "brms", "cli", "cmdstanr", "coda", "dplyr",
  "emmeans", "ggbeeswarm", "ggplot2", "here", "httr2", "janitor",
  "kableExtra", "knitr", "orchaRd", "patchwork", "png", "posterior", "purrr",
  "quarto", "readr", "remotes", "rotl", "scales", "sessioninfo", "stringr", "tibble",
  "tidybayes", "tidyr", "tidyverse"
)
install.packages(setdiff(cran_packages, rownames(installed.packages())))
```

Install `prepR4pcm` from its source repository:

```r
install.packages("remotes")
remotes::install_github("itchyshin/prepR4pcm")
```

Check the complete project list:

```r
source(here::here("Scripts", "00_packages.R"))
```

No message means that the package check passed.

## Step 3 — Install CmdStan

The Bayesian models use CmdStan through `cmdstanr`.

```r
cmdstanr::install_cmdstan(cores = max(1, parallel::detectCores() - 1))
cmdstanr::cmdstan_version()
```

The model scripts use CPU chains. A GPU is not required.

## Step 4 — Choose a route

### Quick route: render saved results

Confirm that the main saved files exist:

```r
required_files <- c(
  "Rdata/data_clean/proceed_clean_filtered.rds",
  "Rdata/effect_sizes/proceed_lnm_safe.rds",
  "Rdata/phylogeny/proceed_A_matrix.rds",
  "Rdata/tables/grid_summary_cache.rds"
)
missing_files <- required_files[!file.exists(required_files)]
if (length(missing_files)) stop("Missing: ", paste(missing_files, collapse = ", "))
```

Then move to Step 6.

### Full route: rebuild analysis files

Run the work in this order:

1. [Prepare the data](01_data_preparation.qmd).
2. [Calculate SAFE lnM](02_lnm_effect_sizes.qmd).
3. [Build the dependency structure](04_dependency_structure.qmd).
4. [Build and align the phylogeny](04b_phylogeny.qmd).
5. [Fit the model grid](05_location_scale_model_grid.qmd).
6. Run the precompute scripts in Step 5 below.
7. [Create the publication figures](08_publication_figures.qmd).

The fitting scripts live in `Scripts/` and run from the project root. Each one
loads the fitting toolchain through `Scripts/00_packages_fit.R` and resolves
every path with `here::here()`, so no path editing is needed:

```bash
Rscript Scripts/remote_rebuild_safe.R
Rscript Scripts/run_model_grid.R
Rscript Scripts/run_sensitivity.R
Rscript Scripts/aggregate_sensitivity.R
```

`remote_rebuild_safe.R` rebuilds the cleaned data and the SAFE lnM effect sizes.
`run_model_grid.R` fits the moderator grid; `run_model_grid_m07plus.R` and the
single-model `run_m*.R` scripts refit individual models. Fitting the full grid
takes many hours and needs roughly 400 MB of disk per model.

The rendered book is read-only. Rebuild large files with the analysis scripts,
not during a normal book render.

## Step 5 — Prepare small files for the book

Every script builds its output paths with `dir_out()`, defined once in
`Scripts/01_paths.R`. It writes to `Rdata/` by default, so a local run
puts files where the book already reads them and no copying is needed. To send
output somewhere else, set `options(pace.out_root = "outputs")` before sourcing
`01_paths.R`.

After fitting models, run these scripts from the project root in this order:

```bash
Rscript Scripts/precompute_grid_summaries.R
Rscript Scripts/precompute_epred_draws.R
Rscript Scripts/precompute_emmeans_contrasts.R
Rscript Scripts/precompute_heterogeneity.R
```

These scripts convert large fitted models into small tables and prediction
files.

A local run needs no copying. If the models were fitted on a separate compute
server that writes to `outputs/`, copy the completed folders across instead:

| Compute server | Web-book project |
|---|---|
| `outputs/epred_draws/` | `Rdata/epred_draws/` |
| `outputs/summaries/` | `Rdata/summaries/` |
| `outputs/tables/` | `Rdata/tables/` |
| `outputs/figures/` | `Rdata/figures/` |

Replace the matching files in `Rdata/` only after every compute script finishes.

::: {.callout-note title="Fitted models are not in the repository"}
`.gitignore` excludes every `.rds` file, so a fresh clone contains the raw data,
all scripts, and the small CSV and figure outputs, but none of the fitted models
or cached draws. Take the quick route only if you already hold a copy of
`Rdata/`; otherwise use the full route and refit.
:::

## Step 6 — Render the web book

```r
quarto::quarto_render()
```

The website is written to `docs/`. Open `docs/index.html` to read the local
copy. Missing saved results cause an error instead of starting a long fit.

## Step 7 — Record the exact environment

### Package inventory

```{r setup}
source(here::here("Scripts", "00_packages.R"))
```

```{r package-inventory}
package_versions <- tibble::tibble(
  package = project_packages,
  version = vapply(project_packages, function(x) {
    if (requireNamespace(x, quietly = TRUE)) {
      as.character(utils::packageVersion(x))
    } else {
      "not installed"
    }
  }, character(1))
)
knitr::kable(package_versions, col.names = c("Package", "Version"))
```

### R and platform

```{r session-info}
sessioninfo::session_info()
```

### CmdStan

```{r cmdstan-version}
tryCatch(
  cat("CmdStan version:", cmdstanr::cmdstan_version(), "\n"),
  error = function(e) cat("CmdStan not available:", conditionMessage(e), "\n")
)
```

### Git commit

```{r git-info}
git_hash <- tryCatch(system("git rev-parse HEAD", intern = TRUE),
                     error = function(e) "git not available")
git_status <- tryCatch(system("git status --short", intern = TRUE),
                       error = function(e) character(0))
cat("Commit hash:", git_hash, "\n\n")
if (length(git_status)) {
  cat("Uncommitted changes:\n", paste(git_status, collapse = "\n"), "\n")
} else {
  cat("Working tree clean.\n")
}
```

### Random seeds

SAFE lnM calculations use `seed = 123`. Model fits use `seed = 123`. The seeds
are set in `Scripts/04_lnm_safe.R` and
`Scripts/08_fit_or_read_model.R`.

## Step 8 — Check saved results

### Effect-size file

```{r es-cache}
es_path <- here::here("Rdata", "effect_sizes", "proceed_lnm_safe.rds")
if (file.exists(es_path)) {
  cat("File:", es_path, "\n")
  cat("Size:", round(file.size(es_path) / 1024^2, 2), "MB\n")
  cat("Modified:", format(file.mtime(es_path)), "\n")
} else {
  cat("Missing. Complete Step 4, item 2.\n")
}
```

### Prediction files used by model chapters

```{r rendered-cache-audit}
rendered_ids <- c("m00", "m01", "m02", "m03", "m04", "m05", "m07", "m08")
expected_n <- c(m00 = 7186L, m01 = 7186L, m02 = 7186L, m03 = 7186L,
                m04 = 7186L, m05 = 7186L, m07 = 7186L, m08 = 7186L)
cache_audit <- purrr::map_dfr(rendered_ids, function(id) {
  path <- here::here("Rdata", "epred_draws", paste0(id, ".rds"))
  if (!file.exists(path)) {
    return(tibble::tibble(model = id, file = "missing", observations = NA_integer_,
                          expected = unname(expected_n[id]), matches = FALSE))
  }
  cache <- readRDS(path)
  summary_text <- paste(cache$summary_txt, collapse = " ")
  match <- regmatches(summary_text,
                      regexec("Number of observations: ([0-9]+)", summary_text))[[1]]
  observed_n <- if (length(match) >= 2) as.integer(match[2]) else NA_integer_
  tibble::tibble(model = id, file = "found", observations = observed_n,
                 expected = unname(expected_n[id]),
                 matches = !is.na(observed_n) && observed_n == unname(expected_n[id]))
})
knitr::kable(cache_audit, caption = "Saved prediction files used by this book.")
```

::: {.callout-tip title="Reproducibility checklist"}
- The package check completes.
- CmdStan reports a version.
- The required saved files are present.
- Every prediction-file row says `found` and `TRUE`.
- `quarto::quarto_render()` writes `docs/index.html`.
- The commit hash and any uncommitted changes are recorded above.
:::
