---
title: "Reproduce this analysis"
---

::: {.workflow}
::: {}
**Input**
The complete project folder.
:::
::: {}
**Action**
Install software, choose a route, and render.
:::
::: {}
**Output**
The web book and an environment record.
:::
::: {}
**Next**
Check every item in the final checklist.
:::
:::

## Before you start {#sec-reproducibility}

This chapter gives the setup steps in order. Run all commands from the project
folder. Use the **quick route** to rebuild the web book from saved results. Use
the **full route** to calculate the effect sizes and fit the models again.

::: {.callout-important title="Use the project folder"}
Open `Magnitude_pace_modern_life.Rproj` before running R code. Do not change the
folder names. The code uses `here::here()` to find files from the project root.
:::

## Step 1 — Get the project

Download the repository from GitHub or clone it with Git:

```bash
git clone https://github.com/Santiago-0rtega/Magnitude_pace_modern_life.git
cd Magnitude_pace_modern_life
```

Keep the `data/`, `Rdata/`, `chapters/`, and `_quarto.yml` paths unchanged.

## Step 2 — Install R packages

The project checks one package list in `chapters/Rscripts/00_packages.R`. The
list includes every non-base R package named in the book and analysis scripts,
including `ggbeeswarm` and `patchwork`.

```r
cran_packages <- c(
  "ape", "bayesplot", "brms", "cli", "cmdstanr", "coda", "dplyr",
  "emmeans", "ggbeeswarm", "ggplot2", "here", "httr2", "janitor",
  "kableExtra", "knitr", "orchaRd", "patchwork", "png", "posterior", "purrr",
  "quarto", "readr", "remotes", "rotl", "scales", "sessioninfo", "stringr", "tibble",
  "tidybayes", "tidyr", "tidyverse"
)
install.packages(setdiff(cran_packages, rownames(installed.packages())))
```

Install `prepR4pcm` from its source repository:

```r
install.packages("remotes")
remotes::install_github("itchyshin/prepR4pcm")
```

Check the complete project list:

```r
source(here::here("chapters", "Rscripts", "00_packages.R"))
```

No message means that the package check passed.

## Step 3 — Install CmdStan

The Bayesian models use CmdStan through `cmdstanr`.

```r
cmdstanr::install_cmdstan(cores = max(1, parallel::detectCores() - 1))
cmdstanr::cmdstan_version()
```

The model scripts use CPU chains. A GPU is not required.

## Step 4 — Choose a route

### Quick route: render saved results

Confirm that the main saved files exist:

```r
required_files <- c(
  "Rdata/data_clean/proceed_clean_filtered.rds",
  "Rdata/effect_sizes/proceed_lnm_safe.rds",
  "Rdata/phylogeny/proceed_A_matrix.rds",
  "Rdata/tables/grid_summary_cache.rds"
)
missing_files <- required_files[!file.exists(required_files)]
if (length(missing_files)) stop("Missing: ", paste(missing_files, collapse = ", "))
```

Then move to Step 6.

### Full route: rebuild analysis files

Run the work in this order:

1. [Prepare the data](01_data_preparation.qmd).
2. [Calculate SAFE lnM](02_lnm_effect_sizes.qmd).
3. [Build the dependency structure](04_dependency_structure.qmd).
4. [Build and align the phylogeny](04b_phylogeny.qmd).
5. [Fit the model grid](05_location_scale_model_grid.qmd).
6. Run the precompute scripts in Step 5 below.
7. [Create the publication figures](08_publication_figures.qmd).

The rendered book is read-only. Rebuild large files with the analysis scripts,
not during a normal book render.

## Step 5 — Prepare small files for the book

The compute server uses `outputs/` as its working output folder. The web book
uses `Rdata/` as its local, read-only mirror. After fitting models, run these
scripts from the compute-project root and in this order:

```bash
Rscript chapters/Rscripts/precompute_grid_summaries.R
Rscript chapters/Rscripts/precompute_epred_draws.R
Rscript chapters/Rscripts/precompute_emmeans_contrasts.R
Rscript chapters/Rscripts/precompute_heterogeneity.R
```

These scripts convert large fitted models into small tables and prediction
files.

Copy the completed folders from the compute server to the matching book paths:

| Compute server | Web-book project |
|---|---|
| `outputs/epred_draws/` | `Rdata/epred_draws/` |
| `outputs/summaries/` | `Rdata/summaries/` |
| `outputs/tables/` | `Rdata/tables/` |
| `outputs/figures/` | `Rdata/figures/` |

Replace the matching files in `Rdata/` only after every compute script finishes.
The model chapters read the copied files; they never read `outputs/`.

## Step 6 — Render the web book

```r
quarto::quarto_render()
```

The website is written to `docs/`. Open `docs/index.html` to read the local
copy. Missing saved results cause an error instead of starting a long fit.

## Step 7 — Record the exact environment

### Package inventory

```{r setup}
source(here::here("chapters", "Rscripts", "00_packages.R"))
```

```{r package-inventory}
package_versions <- tibble::tibble(
  package = project_packages,
  version = vapply(project_packages, function(x) {
    if (requireNamespace(x, quietly = TRUE)) {
      as.character(utils::packageVersion(x))
    } else {
      "not installed"
    }
  }, character(1))
)
knitr::kable(package_versions, col.names = c("Package", "Version"))
```

### R and platform

```{r session-info}
sessioninfo::session_info()
```

### CmdStan

```{r cmdstan-version}
tryCatch(
  cat("CmdStan version:", cmdstanr::cmdstan_version(), "\n"),
  error = function(e) cat("CmdStan not available:", conditionMessage(e), "\n")
)
```

### Git commit

```{r git-info}
git_hash <- tryCatch(system("git rev-parse HEAD", intern = TRUE),
                     error = function(e) "git not available")
git_status <- tryCatch(system("git status --short", intern = TRUE),
                       error = function(e) character(0))
cat("Commit hash:", git_hash, "\n\n")
if (length(git_status)) {
  cat("Uncommitted changes:\n", paste(git_status, collapse = "\n"), "\n")
} else {
  cat("Working tree clean.\n")
}
```

### Random seeds

SAFE lnM calculations use `seed = 123`. Model fits use `seed = 123`. The seeds
are set in `chapters/Rscripts/04_lnm_safe.R` and
`chapters/Rscripts/08_fit_or_read_model.R`.

## Step 8 — Check saved results

### Effect-size file

```{r es-cache}
es_path <- here::here("Rdata", "effect_sizes", "proceed_lnm_safe.rds")
if (file.exists(es_path)) {
  cat("File:", es_path, "\n")
  cat("Size:", round(file.size(es_path) / 1024^2, 2), "MB\n")
  cat("Modified:", format(file.mtime(es_path)), "\n")
} else {
  cat("Missing. Complete Step 4, item 2.\n")
}
```

### Prediction files used by model chapters

```{r rendered-cache-audit}
rendered_ids <- c("m00", "m01", "m02", "m03", "m04", "m05", "m07", "m08")
expected_n <- c(m00 = 7186L, m01 = 7186L, m02 = 7186L, m03 = 7186L,
                m04 = 7186L, m05 = 7186L, m07 = 7186L, m08 = 7186L)
cache_audit <- purrr::map_dfr(rendered_ids, function(id) {
  path <- here::here("Rdata", "epred_draws", paste0(id, ".rds"))
  if (!file.exists(path)) {
    return(tibble::tibble(model = id, file = "missing", observations = NA_integer_,
                          expected = unname(expected_n[id]), matches = FALSE))
  }
  cache <- readRDS(path)
  summary_text <- paste(cache$summary_txt, collapse = " ")
  match <- regmatches(summary_text,
                      regexec("Number of observations: ([0-9]+)", summary_text))[[1]]
  observed_n <- if (length(match) >= 2) as.integer(match[2]) else NA_integer_
  tibble::tibble(model = id, file = "found", observations = observed_n,
                 expected = unname(expected_n[id]),
                 matches = !is.na(observed_n) && observed_n == unname(expected_n[id]))
})
knitr::kable(cache_audit, caption = "Saved prediction files used by this book.")
```

::: {.callout-tip title="Reproducibility checklist"}
- The package check completes.
- CmdStan reports a version.
- The required saved files are present.
- Every prediction-file row says `found` and `TRUE`.
- `quarto::quarto_render()` writes `docs/index.html`.
- The commit hash and any uncommitted changes are recorded above.
:::
