---
title: "m00 — Intercept-only baseline"
---

```{r setup, include=FALSE}
source(here::here("R", "14_epred_cache.R"))
cache_dir <- here::here("outputs", "epred_draws")
cache     <- read_epred_cache("m00", cache_dir)
```

The intercept-only model estimates the overall mean lnM and the baseline residual heterogeneity (sigma), with no moderators — the reference against which every moderator model is compared.

## Model summary

`summary()` is read from the pre-computed cache (`outputs/epred_draws/m00.rds`) — the full `brms` fit is never loaded at render.

```{r summary}
print_cache_summary(cache)
```

## Orchard-like model figure

The orchard-style figure is read from `outputs/figures/publication/orchard/`. The complete editable code used to build this plot is shown below.

```{r figure, fig.width=8, fig.height=9}
knitr::include_graphics(
  here::here("outputs", "figures", "publication", "orchard", "m00_orchard_combined.png")
)
```
### Editable plotting code

```{r build-orchard-m00, eval=FALSE}
library(here)
library(tidyverse)
library(brms)
library(ggbeeswarm)
library(patchwork)

out_dir <- here::here("outputs", "figures", "publication", "orchard")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

dat_es <- readRDS(here::here("outputs", "effect_sizes", "proceed_lnm_safe.rds"))
fit <- readRDS(here::here("outputs", "models", "m00_ls_intercept_only.rds"))

theme_orchard <- function() {
  ggplot2::theme_classic(base_size = 13) +
    ggplot2::theme(
      axis.text.y        = ggplot2::element_text(size = 11),
      axis.title.x       = ggplot2::element_text(size = 12),
      axis.ticks.y       = ggplot2::element_blank(),
      panel.grid.major.x = ggplot2::element_line(colour = "grey92"),
      legend.position    = "bottom",
      legend.title       = ggplot2::element_text(size = 10),
      plot.title         = ggplot2::element_text(size = 13, face = "bold"),
      plot.caption       = ggplot2::element_text(size = 9, colour = "grey50")
    )
}

save_plot <- function(p, stem, width = 9, height = 6) {
  ggplot2::ggsave(file.path(out_dir, paste0(stem, ".pdf")), p,
                  width = width, height = height, device = cairo_pdf)
  ggplot2::ggsave(file.path(out_dir, paste0(stem, ".png")), p,
                  width = width, height = height, dpi = 300, type = "cairo")
}

raw <- dat_es |>
  dplyr::mutate(precision = 1 / sqrt(vi_lnM_safe))

fe <- brms::fixef(fit)
int_est <- fe["Intercept", "Estimate"]
int_lo  <- fe["Intercept", "Q2.5"]
int_hi  <- fe["Intercept", "Q97.5"]
sig_est <- fe["sigma_Intercept", "Estimate"]
sig_lo  <- fe["sigma_Intercept", "Q2.5"]
sig_hi  <- fe["sigma_Intercept", "Q97.5"]

vc <- brms::VarCorr(fit)
caption_txt <- sprintf(
  "n = %d effect sizes | study SD = %.2f | phylogeny SD = %.2f",
  nrow(raw), vc$ref_id$sd["Intercept", "Estimate"],
  vc$sp_ncbi$sd["Intercept", "Estimate"]
)

p_loc <- ggplot2::ggplot() +
  ggplot2::geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  ggbeeswarm::geom_quasirandom(
    data = raw,
    ggplot2::aes(x = yi_lnM_safe, y = 1, size = precision),
    alpha = 0.20, shape = 21, fill = "#88CCEE", colour = "#0072B2",
    groupOnX = FALSE
  ) +
  ggplot2::geom_linerange(
    ggplot2::aes(y = 1, xmin = int_lo, xmax = int_hi),
    linewidth = 1.5, colour = "grey15"
  ) +
  ggplot2::geom_point(
    ggplot2::aes(x = int_est, y = 1),
    size = 5, shape = 21, fill = "white", colour = "grey10", stroke = 1.2
  ) +
  ggplot2::scale_size_continuous(name = "Precision (1/SE)", range = c(0.3, 4)) +
  ggplot2::scale_y_continuous(breaks = NULL) +
  ggplot2::labs(
    x = "Location effect (lnM)",
    y = NULL,
    title = "Location -- Overall baseline (m00)",
    caption = caption_txt
  ) +
  theme_orchard()

raw_sig <- raw |>
  dplyr::mutate(log_abs_resid = log(abs(yi_lnM_safe - int_est) + 1e-6))

p_scl <- ggplot2::ggplot() +
  ggplot2::geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  ggbeeswarm::geom_quasirandom(
    data = raw_sig,
    ggplot2::aes(x = log_abs_resid, y = 1, size = precision),
    alpha = 0.20, shape = 21, fill = "#88CCEE", colour = "#0072B2",
    groupOnX = FALSE
  ) +
  ggplot2::geom_linerange(
    ggplot2::aes(y = 1, xmin = sig_lo, xmax = sig_hi),
    linewidth = 3, colour = "white"
  ) +
  ggplot2::geom_linerange(
    ggplot2::aes(y = 1, xmin = sig_lo, xmax = sig_hi),
    linewidth = 1.5, colour = "grey15"
  ) +
  ggplot2::geom_point(
    ggplot2::aes(x = sig_est, y = 1),
    size = 5, shape = 21, fill = "white", colour = "grey10", stroke = 1.2
  ) +
  ggplot2::scale_size_continuous(name = "Precision (1/SE)", range = c(0.3, 4)) +
  ggplot2::scale_y_continuous(breaks = NULL) +
  ggplot2::labs(
    x = "Residual lnM",
    y = NULL,
    title = "Scale -- Overall baseline (m00)",
    caption = sprintf(
      "sigma_Intercept = %.3f [%.3f, %.3f] | exp(sigma) = %.3f",
      sig_est, sig_lo, sig_hi, exp(sig_est)
    )
  ) +
  theme_orchard()

p_comb <- patchwork::wrap_plots(p_loc, p_scl, ncol = 1)

save_plot(p_loc,  "m00_orchard_location", width = 9, height = 4)
save_plot(p_scl,  "m00_orchard_scale",    width = 9, height = 4)
save_plot(p_comb, "m00_orchard_combined", width = 9, height = 8)
p_comb
```
