---
title: "m04 — Elapsed time (log₁₀ generations)"
---

```{r setup, include=FALSE}
source(here::here("R", "14_epred_cache.R"))
cache_dir <- here::here("outputs", "epred_draws")
cache     <- read_epred_cache("m04", cache_dir)
```

Continuous moderator: divergence as a function of elapsed time in generations (log₁₀).

## Model summary

`summary()` is read from the pre-computed cache (`outputs/epred_draws/m04.rds`) — the full `brms` fit is never loaded at render.

```{r summary}
print_cache_summary(cache)
```

## Orchard-like model figure

The orchard-style figure is read from `outputs/figures/publication/orchard/`. The full figure-generation code is kept in `R/16_plot_orchard_style_v2.R`.

```{r figure, fig.width=8, fig.height=9}
knitr::include_graphics(
  here::here("outputs", "figures", "publication", "orchard", "m04_orchard_combined.png")
)
```
### Regenerate this figure

```{r regenerate-orchard-m04, eval=FALSE}
Sys.setenv(ORCHARD_MODEL_IDS = "m04")
source(here::here("R", "16_plot_orchard_style_v2.R"))
```
