---
title: "m04 — Elapsed time (log₁₀ generations)"
---

```{r setup, include=FALSE}
source(here::here("R", "14_epred_cache.R"))
cache_dir <- here::here("outputs", "epred_draws")
cache     <- read_epred_cache("m04", cache_dir)
```

Continuous moderator: divergence as a function of elapsed time in generations (log₁₀).

## Model summary

`summary()` is read from the pre-computed cache (`outputs/epred_draws/m04.rds`) — the full `brms` fit is never loaded at render.

```{r summary}
print_cache_summary(cache)
```

## Orchard-like model figure

The orchard-style figure is read from `outputs/figures/publication/orchard/`. The complete editable code used to build this plot is shown below.

```{r figure, fig.width=8, fig.height=9}
knitr::include_graphics(
  here::here("outputs", "figures", "publication", "orchard", "m04_orchard_combined.png")
)
```
### Editable plotting code

```{r build-orchard-m04, eval=FALSE}
library(here)
library(tidyverse)
library(brms)
library(tidybayes)
library(patchwork)

model_id <- "m04"
model_file <- "m04_ls_log10_generations"
moderator <- "log10_generations"
moderator_label <- "Elapsed time (log10 generations)"

out_dir <- here::here("outputs", "figures", "publication", "orchard")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

dat_es <- readRDS(here::here("outputs", "effect_sizes", "proceed_lnm_safe.rds"))
fit <- readRDS(here::here("outputs", "models", paste0(model_file, ".rds")))

theme_orchard <- function() {
  ggplot2::theme_classic(base_size = 13) +
    ggplot2::theme(
      axis.text.y        = ggplot2::element_text(size = 11),
      axis.title.x       = ggplot2::element_text(size = 12),
      axis.ticks.y       = ggplot2::element_blank(),
      panel.grid.major.x = ggplot2::element_line(colour = "grey92"),
      legend.position    = "bottom",
      legend.title       = ggplot2::element_text(size = 10),
      plot.title         = ggplot2::element_text(size = 13, face = "bold"),
      plot.caption       = ggplot2::element_text(size = 9, colour = "grey50")
    )
}

save_plot <- function(p, stem, width = 8, height = 6) {
  ggplot2::ggsave(file.path(out_dir, paste0(stem, ".pdf")), p,
                  width = width, height = height, device = cairo_pdf)
  ggplot2::ggsave(file.path(out_dir, paste0(stem, ".png")), p,
                  width = width, height = height, dpi = 300, type = "cairo")
}

raw <- dat_es[!is.na(dat_es[[moderator]]), ] |>
  dplyr::mutate(precision = 1 / sqrt(vi_lnM_safe))

rng <- range(raw[[moderator]], na.rm = TRUE)
nd <- data.frame(x = seq(rng[1], rng[2], length.out = 100))
names(nd) <- moderator
nd[["es_id_model"]] <- NA
nd[["ref_id"]] <- NA
nd[["sp_ncbi"]] <- NA

pred <- tidybayes::add_epred_draws(nd, fit, re_formula = NA, ndraws = 500) |>
  dplyr::group_by(.data[[moderator]]) |>
  dplyr::summarise(
    estimate = median(.epred),
    lowerCL = quantile(.epred, 0.025),
    upperCL = quantile(.epred, 0.975),
    .groups = "drop"
  )

pred_sig <- tidybayes::add_epred_draws(
  nd, fit, dpar = "sigma", re_formula = NA, ndraws = 500
) |>
  dplyr::group_by(.data[[moderator]]) |>
  dplyr::summarise(
    estimate = median(.epred),
    lowerCL = quantile(.epred, 0.025),
    upperCL = quantile(.epred, 0.975),
    .groups = "drop"
  )

raw_pred_loc <- approx(
  pred[[moderator]], pred$estimate,
  xout = raw[[moderator]], rule = 2
)$y

raw_cont_sig <- raw |>
  dplyr::mutate(log_abs_resid = log(abs(yi_lnM_safe - raw_pred_loc) + 1e-6))

p_loc <- ggplot2::ggplot() +
  ggplot2::geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  ggplot2::geom_point(
    data = raw,
    ggplot2::aes(x = .data[[moderator]], y = yi_lnM_safe, size = precision),
    alpha = 0.22, shape = 21, fill = "#88CCEE", colour = "#0072B2"
  ) +
  ggplot2::geom_ribbon(
    data = pred,
    ggplot2::aes(x = .data[[moderator]], ymin = lowerCL, ymax = upperCL),
    alpha = 0.35, fill = "#0072B2"
  ) +
  ggplot2::geom_line(
    data = pred,
    ggplot2::aes(x = .data[[moderator]], y = estimate),
    linewidth = 1.1, colour = "#0072B2"
  ) +
  ggplot2::scale_size_continuous(name = "Precision (1/SE)", range = c(0.3, 4)) +
  ggplot2::labs(
    x = moderator_label,
    y = "lnM",
    title = paste("Location --", moderator_label)
  ) +
  theme_orchard()

p_scl <- ggplot2::ggplot() +
  ggplot2::geom_point(
    data = raw_cont_sig,
    ggplot2::aes(x = .data[[moderator]], y = log_abs_resid, size = precision),
    alpha = 0.20, shape = 21, fill = "#88CCEE", colour = "#0072B2"
  ) +
  ggplot2::geom_ribbon(
    data = pred_sig,
    ggplot2::aes(x = .data[[moderator]], ymin = lowerCL, ymax = upperCL),
    alpha = 0.35, fill = "grey40"
  ) +
  ggplot2::geom_line(
    data = pred_sig,
    ggplot2::aes(x = .data[[moderator]], y = estimate),
    linewidth = 1.1, colour = "grey15"
  ) +
  ggplot2::scale_size_continuous(name = "Precision (1/SE)", range = c(0.3, 4)) +
  ggplot2::labs(
    x = moderator_label,
    y = "Residual lnM",
    title = paste("Scale --", moderator_label),
    caption = "Higher sigma = greater residual heterogeneity."
  ) +
  theme_orchard()

p_comb <- patchwork::wrap_plots(p_loc, p_scl, ncol = 1)

save_plot(p_loc, paste0(model_id, "_orchard_location"), width = 8, height = 5)
save_plot(p_scl, paste0(model_id, "_orchard_scale"), width = 8, height = 4)
save_plot(p_comb, paste0(model_id, "_orchard_combined"), width = 8, height = 9)
p_comb
```
