---
title: "Protocol deviation — moderators adjusted for elapsed time"
---

## Recorded change

This chapter was **not** part of the pre-registered protocol. It refits the primary moderators
(`disturbance`, `design`, `trait_type`, `genphen`) with elapsed time added as an additive covariate.

::: {.callout-note title="What changed—and what did not"}
**Changed:** for each of the four primary moderators, we fitted an additive
location-scale model that adds elapsed time (`log10_years`, and separately
`log10_generations`) as a second main-effect term, with **no interaction**.
This adjusts each moderator's estimate for the average elapsed-time
difference between its levels, without letting the time slope itself vary
by level.

**Unchanged:** the dataset, the moderator definitions, the `ref_id` and
phylogenetic random effects, the priors, and the MCMC settings all match the
primary single-moderator models.
:::

This produces eight models: four moderators × two elapsed-time scales.

## Moderator–time relationships

The plots below show how elapsed time is
distributed across the levels of each moderator.

```{r setup}
source(here::here("Scripts", "00_packages.R"))

dat_es <- readRDS(here::here("Rdata", "effect_sizes", "proceed_lnm_safe.rds"))

moderator_labels <- c(
  disturbance = "Disturbance context",
  design      = "Comparison design",
  trait_type  = "Trait type",
  genphen     = "Phenotypic vs. genetic"
)

theme_deviation <- function(base_size = 13) {
  ggplot2::theme_classic(base_size = base_size) +
    ggplot2::theme(
      strip.background = ggplot2::element_blank(),
      strip.text        = ggplot2::element_text(face = "bold"),
      axis.text.x       = ggplot2::element_text(angle = 30, hjust = 1),
      panel.grid.major.y = ggplot2::element_line(colour = "grey92"),
      plot.title        = ggplot2::element_text(face = "bold")
    )
}

make_time_panel <- function(dat, timevar, timelabel) {
  long_dat <- purrr::map_dfr(names(moderator_labels), function(m) {
    dat |>
      dplyr::filter(!is.na(.data[[m]]), !is.na(.data[[timevar]])) |>
      dplyr::transmute(
        moderator_label = moderator_labels[[m]],
        level = as.character(.data[[m]]),
        time_value = .data[[timevar]]
      )
  })

  ggplot2::ggplot(long_dat, ggplot2::aes(x = level, y = time_value)) +
    ggplot2::geom_boxplot(fill = "grey85", outlier.alpha = 0.35, width = 0.6) +
    ggplot2::facet_wrap(~moderator_label, scales = "free_x", nrow = 2) +
    ggplot2::labs(x = NULL, y = timelabel) +
    theme_deviation()
}
```

```{r plot-years, fig.width=11, fig.height=8, fig.cap="Elapsed years (log10 scale) across the levels of each primary moderator."}
make_time_panel(dat_es, "log10_years", expression(log[10]("elapsed years")))
```

```{r plot-generations, fig.width=11, fig.height=8, fig.cap="Elapsed generations (log10 scale) across the levels of each primary moderator."}
make_time_panel(dat_es, "log10_generations", expression(log[10]("elapsed generations")))
```

## Model specification

Models were fitted on a remote server, not while rendering this book, using the same
priors and MCMC settings as the primary models: 4 chains × 4,000 iterations
(2,000 warmup), `adapt_delta = 0.97`, `max_treedepth = 15`, `cmdstanr` backend.

```{r how-fit, eval=FALSE}
loc_formula <- as.formula(paste0(
  "yi_lnM_safe ~ ", moderator, " + ", timevar,
  " + (1 | ref_id) + (1 | gr(sp_ncbi_canonical, cov = A))",
  " + (1 | gr(es_id_model, cov = V))"
))
scl_formula <- as.formula(paste0("sigma ~ ", moderator, " + ", timevar))

formula_add <- brms::bf(loc_formula, scl_formula)
fit_add <- brms::brm(
  formula = formula_add, data = dat_model, data2 = list(A = A_mod, V = V),
  prior = priors_add, chains = 4, iter = 4000, warmup = 2000, cores = 4,
  backend = "cmdstanr", control = list(adapt_delta = 0.97, max_treedepth = 15)
)
```

## Convergence diagnostics

```{r diagnostics}
diag_tbl <- readr::read_csv(
  here::here("Rdata", "tables", "additive_diagnostics.csv"),
  show_col_types = FALSE
)
knitr::kable(
  diag_tbl,
  col.names = c("Model", "Moderator", "Elapsed-time variable", "Max Rhat",
               "Min Bulk ESS", "Min Tail ESS", "Divergent transitions",
               "Max-treedepth hits"),
  digits = 3,
  caption = "Convergence diagnostics for the eight additive (moderator + time) models."
)
```

## Model summaries

```{r summaries-setup, include=FALSE}
additive_summaries <- readRDS(here::here("Rdata", "summaries", "additive_summaries.rds"))
```

::: {.panel-tabset}

### disturbance + years

```{r summary-disturbance-years, echo=FALSE}
writeLines(additive_summaries[["disturbance_plus_log10_years"]])
```

### disturbance + generations

```{r summary-disturbance-generations, echo=FALSE}
writeLines(additive_summaries[["disturbance_plus_log10_generations"]])
```

### design + years

```{r summary-design-years, echo=FALSE}
writeLines(additive_summaries[["design_plus_log10_years"]])
```

### design + generations

```{r summary-design-generations, echo=FALSE}
writeLines(additive_summaries[["design_plus_log10_generations"]])
```

### trait_type + years

```{r summary-trait-years, echo=FALSE}
writeLines(additive_summaries[["trait_type_plus_log10_years"]])
```

### trait_type + generations

```{r summary-trait-generations, echo=FALSE}
writeLines(additive_summaries[["trait_type_plus_log10_generations"]])
```

### genphen + years

```{r summary-genphen-years, echo=FALSE}
writeLines(additive_summaries[["genphen_plus_log10_years"]])
```

### genphen + generations

```{r summary-genphen-generations, echo=FALSE}
writeLines(additive_summaries[["genphen_plus_log10_generations"]])
```

:::

## Orchard-style figures

Each figure shows the focal moderator's location (top) and scale (bottom)
estimates from its additive model, with elapsed time held at its dataset
mean. Points and beeswarm markers are on the same scale as the primary
single-moderator orchard figures (see the model chapters); only the added
time term differs.

::: {.panel-tabset}

### Disturbance

```{r orchard-disturbance-years, fig.width=13, fig.height=12}
knitr::include_graphics(here::here("Rdata", "figures", "publication",
                                   "orchard_additive",
                                   "disturbance_plus_log10_years_orchard_combined.png"))
```

```{r orchard-disturbance-generations, fig.width=13, fig.height=12}
knitr::include_graphics(here::here("Rdata", "figures", "publication",
                                   "orchard_additive",
                                   "disturbance_plus_log10_generations_orchard_combined.png"))
```

### Design

```{r orchard-design-years, fig.width=9, fig.height=10}
knitr::include_graphics(here::here("Rdata", "figures", "publication",
                                   "orchard_additive",
                                   "design_plus_log10_years_orchard_combined.png"))
```

```{r orchard-design-generations, fig.width=9, fig.height=10}
knitr::include_graphics(here::here("Rdata", "figures", "publication",
                                   "orchard_additive",
                                   "design_plus_log10_generations_orchard_combined.png"))
```

### Trait type

```{r orchard-trait-years, fig.width=13, fig.height=12}
knitr::include_graphics(here::here("Rdata", "figures", "publication",
                                   "orchard_additive",
                                   "trait_type_plus_log10_years_orchard_combined.png"))
```

```{r orchard-trait-generations, fig.width=13, fig.height=12}
knitr::include_graphics(here::here("Rdata", "figures", "publication",
                                   "orchard_additive",
                                   "trait_type_plus_log10_generations_orchard_combined.png"))
```

### Genphen

```{r orchard-genphen-years, fig.width=9, fig.height=10}
knitr::include_graphics(here::here("Rdata", "figures", "publication",
                                   "orchard_additive",
                                   "genphen_plus_log10_years_orchard_combined.png"))
```

```{r orchard-genphen-generations, fig.width=9, fig.height=10}
knitr::include_graphics(here::here("Rdata", "figures", "publication",
                                   "orchard_additive",
                                   "genphen_plus_log10_generations_orchard_combined.png"))
```

:::

::: {.callout-warning title="Read the diagnostics table before the figures above"}
The disturbance and trait_type models did not meet the convergence criteria
used elsewhere in this book (max Rhat ≤ 1.01, bulk and tail ESS ≥ 400, and no
divergent transitions) with either elapsed-time variable; the failures were
most severe with `log10_generations`. Their summaries and
figures are shown here for transparency, but should not be read as reliable
estimates. The design and genphen models converged cleanly.
:::
