---
title: "Sensitivity C — Study-system (sys_id) random effect"
---

```{r setup, include=FALSE}
source(here::here("chapters", "Rscripts", "00_packages.R"))
source(here::here("chapters", "Rscripts", "19_sensitivity_tables.R"))
variant <- "sens3_sysid"
```

## What this tests

Protocol §3.7 (final analysis) and §3.5. Contrasts from the same study system
may be non-independent because they share a population, location, disturbance
history, or research programme. The primary model captures study-level
dependence through `ref_id` and shared ancestry through the phylogeny, but does
**not** include a separate study-system term, because `sys_id`, `ref_id`, and
species identity are strongly associated in PROCEED — most systems map to a
single species and often a single source — so including all three risks poorly
identified variance components. This analysis adds `(1 | sys_id)` to each
included moderator model and reports the outcome explicitly, per the protocol's
commitment to flag weak identifiability rather than interpret confounded
components separately. The full dataset is used.

## How it was fit

Fit on `totoro`, not at render — 4 chains × 4,000 iterations (2,000 warmup),
`adapt_delta = 0.97`, `max_treedepth = 15`, `cmdstanr`.

```{r how-fit, eval=FALSE}
phylo   <- prepare_phylo_and_data(dat_mod, A_full)
formula <- build_ls_formula(moderator, has_phylogeny = phylo$has_phylo,
                            has_sys_id = TRUE)              # + (1 | sys_id)
fit     <- fit_ls_model(
  phylo$dat_model, formula,
  build_ls_priors(formula, phylo$dat_model, phylo$V, A = phylo$A_mod),
  V = phylo$V, A = phylo$A_mod, mcmc_args = default_mcmc_args)
```

## Convergence diagnostics

```{r diagnostics, echo=FALSE}
sens_diagnostics_table(variant)
```

::: {.callout-warning}
The **disturbance** model returned 21 divergent transitions and the **design**
model 7. This is the weak-identifiability signal the protocol anticipated: with
`sys_id` added on top of `ref_id` and the phylogeny, the sampler struggles in
part of the posterior. R̂ (< 1.01) and ESS remain acceptable, but the individual
`sys_id`, `ref_id`, and phylogenetic variance components should **not** be
interpreted separately for these models.
:::

## Reading the tables

Raw location–scale coefficients on treatment coding: the **Intercept** row is the
reference-level value, each other row the difference from it. The matched
**primary** (no `sys_id`) coefficient is shown alongside, and the **Stability**
column reads *consistent* when the two share sign and the same 95%
credible-interval conclusion, **DIFFERS** (highlighted) otherwise. Stability of
the moderator coefficients here indicates the reported effects are not artefacts
of omitting a study-system term.

```{r stability-line, echo=FALSE, results='asis'}
cat("**Overall:** ", sens_stability_summary(variant), "\n")
```

## Results by moderator

```{r results, echo=FALSE, results='asis'}
sens_all_moderator_tables(variant)
```
