---
title: "Sensitivity A — Minimum sample size (N ≥ 40)"
---

```{r setup, include=FALSE}
source(here::here("chapters", "Rscripts", "00_packages.R"))
source(here::here("chapters", "Rscripts", "19_sensitivity_tables.R"))
variant <- "sens1_n40"
sens_dat <- readRDS(here::here("Rdata", "effect_sizes", "proceed_lnm_safe.rds"))
sens_total <- nrow(sens_dat)
sens_n40 <- sum(sens_dat$n_total >= 40, na.rm = TRUE)
```

## What this tests

Protocol §3.7 (first analysis). `ln M` is estimated less stably when the two
groups being compared are small, because the plug-in sampling variance can
become unstable when between-group variation is small relative to within-group
variation. This sensitivity analysis refits every included moderator model
after restricting the data
to contrasts with total sample size $n_1 + n_2 \geq 40$ — **`r format(sens_n40, big.mark = ",")` of `r format(sens_total, big.mark = ",")`
contrasts**. All other model structure (phylogeny, `ref_id`, priors, MCMC
settings) is identical to the primary fit.

## How it was fit

Fit on the compute server (`totoro`), not at render. Each model uses 4 chains ×
4,000 iterations (2,000 warmup), `adapt_delta = 0.97`, `max_treedepth = 15`,
`cmdstanr`. The book reads the cached summary tables only.

```{r how-fit, eval=FALSE}
dat_n40 <- dat_es |> dplyr::filter(n_total >= 40)
phylo   <- prepare_phylo_and_data(dat_mod, A_full)          # phylogeny retained
formula <- build_ls_formula(moderator, has_phylogeny = phylo$has_phylo)
fit     <- fit_ls_model(
  phylo$dat_model, formula,
  build_ls_priors(formula, phylo$dat_model, phylo$V, A = phylo$A_mod),
  V = phylo$V, A = phylo$A_mod, mcmc_args = default_mcmc_args)
```

## Convergence diagnostics

```{r diagnostics, echo=FALSE}
sens_diagnostics_table(variant)
```

::: {.callout-warning}
`r sens_diagnostic_flags(variant)`
:::

## Reading the tables

Each moderator is shown as a raw location–scale coefficient table on the model's
default treatment coding: the **Intercept** row is the reference-level value and
each other row is the difference from that reference. The matched **primary**
coefficient is shown alongside. The **Stability** column compares the two: it
reads *consistent* when the sensitivity and primary coefficients share sign and
the same 95% credible-interval conclusion (both exclude or both include zero),
and **DIFFERS** otherwise.

```{r stability-line, echo=FALSE, results='asis'}
cat("**Overall:** ", sens_stability_summary(variant), "\n")
```

## Results by moderator

```{r results, echo=FALSE, results='asis'}
sens_all_moderator_tables(variant)
```
