---
title: "Sensitivity B — No phylogenetic random effect"
---

```{r setup, include=FALSE}
source(here::here("chapters", "Rscripts", "00_packages.R"))
source(here::here("chapters", "Rscripts", "19_sensitivity_tables.R"))
variant <- "sens2_nophylo"
```

## What this tests

The primary models place a phylogenetic random effect
`(1 | gr(sp_ncbi, cov = A))` on every species-level intercept, using a
correlation matrix `A` derived from an Open Tree of Life scaffold. That scaffold
is a pragmatic comparative tree, not a fully resolved, time-calibrated phylogeny
(protocol §3.6): its topology is assembled from multiple sources, unresolved
nodes are resolved algorithmically, and branch lengths are assigned by Grafen's
method. This analysis refits each included moderator model with the phylogenetic
term **removed**, keeping the `ref_id` study effect and everything else fixed.
The full dataset is used (no sample-size restriction).

## How it was fit

Fit on `totoro`, not at render — 4 chains × 4,000 iterations (2,000 warmup),
`adapt_delta = 0.97`, `max_treedepth = 15`, `cmdstanr`.

```{r how-fit, eval=FALSE}
formula <- build_ls_formula(moderator, has_phylogeny = FALSE)   # drop (1 | gr(sp_ncbi, cov = A))
V       <- diag(dat_mod$vi_lnM_safe)
fit     <- fit_ls_model(
  dat_mod, formula,
  build_ls_priors(formula, dat_mod, V, A = NULL),
  V = V, A = NULL, mcmc_args = default_mcmc_args)
```

## Convergence diagnostics

```{r diagnostics, echo=FALSE}
sens_diagnostics_table(variant)
```

`r sens_diagnostic_flags(variant)`

## Reading the tables

Raw location–scale coefficients on treatment coding: the **Intercept** row is the
reference-level value, each other row the difference from it. The matched
**primary** (phylogeny-included) coefficient is shown alongside, and the
**Stability** column reads *consistent* when the two share sign and the same 95%
credible-interval conclusion, **DIFFERS** otherwise.

```{r stability-line, echo=FALSE, results='asis'}
cat("**Overall:** ", sens_stability_summary(variant), "\n")
```

## Results by moderator

```{r results, echo=FALSE, results='asis'}
sens_all_moderator_tables(variant)
```
