---
title: "Sensitivity B — No phylogenetic random effect"
---

```{r setup, include=FALSE}
source(here::here("chapters", "Rscripts", "00_packages.R"))
source(here::here("chapters", "Rscripts", "19_sensitivity_tables.R"))
variant <- "sens2_nophylo"
```

## Recorded change

The primary models place a phylogenetic random effect
`(1 | gr(sp_ncbi, cov = A))` on every species-level intercept, using a
correlation matrix `A` derived from an Open Tree of Life scaffold. That scaffold
is a pragmatic comparative tree, not a fully resolved, time-calibrated phylogeny
(protocol §3.6): its topology is assembled from multiple sources, unresolved
nodes are resolved algorithmically, and branch lengths are assigned by Grafen's
method.

::: {.callout-note title="What changed—and what did not"}
**Changed:** the phylogenetic random effect was removed.

**Unchanged:** the full dataset, moderator definitions, `ref_id` effect, priors,
and MCMC settings match the primary models.
:::

## Model specification

Models were fit on `totoro`, not while rendering this book: 4 chains × 4,000
iterations (2,000 warmup), `adapt_delta = 0.97`, `max_treedepth = 15`, and the
`cmdstanr` backend.

```{r how-fit, eval=FALSE}
formula <- build_ls_formula(moderator, has_phylogeny = FALSE)
V       <- diag(dat_mod$vi_lnM_safe)
fit     <- fit_ls_model(
  dat_mod, formula,
  build_ls_priors(formula, dat_mod, V, A = NULL),
  V = V, A = NULL, mcmc_args = default_mcmc_args)
```

## Convergence diagnostics

```{r diagnostics, echo=FALSE}
sens_diagnostics_table(variant)
```

::: {.callout-warning}
`r sens_diagnostic_flags(variant)`
:::

## Overall comparison

The comparison label uses a fixed rule: **DIFFERS** means that the posterior
median changed sign or that the 95% credible interval changed between including
and excluding zero. *Consistent* means neither occurred; it does not mean the
two estimates are identical. Intercepts are omitted from the overall count.

```{r stability-line, echo=FALSE, results='asis'}
cat("**Overall:** ", sens_stability_summary(variant), "\n")
```

## Paired estimates by moderator

Each table reports raw treatment-coded coefficients for the location
(mean `ln M`) and scale (log-SD) submodels. See the
[sensitivity overview](../07_sensitivity_analyses.qmd#how-to-read-the-detailed-results)
for column definitions.

```{r results, echo=FALSE, results='asis'}
sens_all_moderator_tables(variant)
```
