---
title: "Sensitivity A — Minimum sample size (N ≥ 40)"
---

```{r setup, include=FALSE}
source(here::here("Scripts", "00_packages.R"))
source(here::here("Scripts", "19_sensitivity_tables.R"))
variant <- "sens1_n40"
sens_dat <- readRDS(here::here("Rdata", "effect_sizes", "proceed_lnm_safe.rds"))
sens_total <- nrow(sens_dat)
sens_n40 <- sum(sens_dat$n_total >= 40, na.rm = TRUE)
```

## Recorded change

Following protocol §3.7, we retained only contrasts with total sample
size $n_1 + n_2 \geq 40$.

::: {.callout-note title="What changed—and what did not"}
**Changed:** the analysis includes **`r format(sens_n40, big.mark = ",")` of
`r format(sens_total, big.mark = ",")` contrasts** after applying the sample-size
threshold.

**Unchanged:** the moderator definitions, phylogenetic effect, `ref_id` effect,
priors, and MCMC settings match the primary models.
:::

## Model specification

Models were fit on a remote server, not while rendering this
book. Each model uses 4 chains × 4,000 iterations (2,000 warmup),
`adapt_delta = 0.97`, `max_treedepth = 15`, and the `cmdstanr` backend. This
chapter reads cached summary tables only.

```{r how-fit, eval=FALSE}
dat_n40 <- dat_es |> dplyr::filter(n_total >= 40)
phylo   <- prepare_phylo_and_data(dat_mod, A_full)
formula <- build_ls_formula(moderator, has_phylogeny = phylo$has_phylo)
fit     <- fit_ls_model(
  phylo$dat_model, formula,
  build_ls_priors(formula, phylo$dat_model, phylo$V, A = phylo$A_mod),
  V = phylo$V, A = phylo$A_mod, mcmc_args = default_mcmc_args)
```

## Convergence diagnostics

Six moderator models are reported (m01–m05, m07). The environmental-change
model (m08) is excluded because its primary fit did not converge. Convergence
criteria: max $\hat{R} \leq 1.01$, bulk and tail ESS $\geq 400$, and no
divergent transitions.

```{r diagnostics}
sens_diagnostics_table(variant)
```

::: {.callout-warning}
`r sens_diagnostic_flags(variant)`
:::

## Overall comparison

The comparison label uses a fixed rule: **DIFFERS** means that the posterior
median changed sign or that the 95% credible interval changed between including
and excluding zero. *Consistent* means neither occurred; it does not mean the
two estimates are identical. Intercepts are omitted from the overall count.

```{r stability-line, results='asis'}
cat("**Overall:** ", sens_stability_summary(variant), "\n")
```

## Paired estimates by moderator

Each table reports raw treatment-coded coefficients for the location
(mean `ln M`) and scale (log-SD) submodels. See the
[sensitivity overview](../07_sensitivity_analyses.qmd#output-columns)
for column definitions.

```{r results, results='asis'}
sens_all_moderator_tables(variant)
```
