---
title: "Sensitivity C — Study-system (sys_id) random effect"
---

```{r setup, include=FALSE}
source(here::here("Scripts", "00_packages.R"))
source(here::here("Scripts", "19_sensitivity_tables.R"))
variant <- "sens3_sysid"
```

## Recorded change

Under protocol §§3.5 and 3.7, contrasts from the same study system may be
non-independent because they share a population, location, disturbance history,
or research programme. The primary model captures study-level dependence
through `ref_id` and shared ancestry through the phylogeny, but does not include
a separate study-system term. In PROCEED, `sys_id`, `ref_id`, and species
identity are strongly associated. Most systems map to one species and often one
source, so fitting all three grouping structures can make their variance
components difficult to distinguish.

::: {.callout-note title="What changed—and what did not"}
**Changed:** a study-system random intercept, `(1 | sys_id)`, was added.

**Unchanged:** the full dataset, moderator definitions, phylogenetic effect,
`ref_id` effect, priors, and MCMC settings match the primary models.

The comparison uses the fixed-effect tables. The three variance components are
not reported separately.
:::

## Model specification

Models were fit on the labs server (`totoro`), not while rendering this book: 4 chains × 4,000
iterations (2,000 warmup), `adapt_delta = 0.97`, `max_treedepth = 15`, and the
`cmdstanr` backend.

```{r how-fit, eval=FALSE}
phylo   <- prepare_phylo_and_data(dat_mod, A_full)
formula <- build_ls_formula(moderator, has_phylogeny = phylo$has_phylo,
                            has_sys_id = TRUE)
fit     <- fit_ls_model(
  phylo$dat_model, formula,
  build_ls_priors(formula, phylo$dat_model, phylo$V, A = phylo$A_mod),
  V = phylo$V, A = phylo$A_mod, mcmc_args = default_mcmc_args)
```

## Convergence diagnostics

Six moderator models are reported (m01–m05, m07). The environmental-change
model (m08) is excluded because its primary fit did not converge. Convergence
criteria: max $\hat{R} \leq 1.01$, bulk and tail ESS $\geq 400$, and no
divergent transitions.

```{r diagnostics}
sens_diagnostics_table(variant)
```

::: {.callout-warning}
`r sens_diagnostic_flags(variant)`
:::

## Overall comparison

The comparison label uses a fixed rule: **DIFFERS** means that the posterior
median changed sign or that the 95% credible interval changed between including
and excluding zero. *Consistent* means neither occurred; it does not mean the
two estimates are identical. Intercepts are omitted from the overall count.

```{r stability-line, results='asis'}
cat("**Overall:** ", sens_stability_summary(variant), "\n")
```

## Paired estimates by moderator

Each table reports raw treatment-coded coefficients for the location
(mean `ln M`) and scale (log-SD) submodels. See the
[sensitivity overview](../07_sensitivity_analyses.qmd#output-columns)
for column definitions.

```{r results, results='asis'}
sens_all_moderator_tables(variant)
```
